# Configuration

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Configuring the workflow
After creating local config files during [Installation](installation.qmd), adapt the workflow by editing the reaction, dataset, aggregation, and optional module config files. These files are the main public interface for changing what the workflow runs.

The terminology is:

- **reaction**: one `cellranger-arc count` output directory from a single 10x reaction.
- **dataset**: a named group of reactions that share reaction-level settings, such as reference genome and QC filters.
- **aggregation**: a configured joint analysis of one or more reactions.

YAML entries are divided into two blocks:

1) **`mandatory`**. Parameters that must be set for the relevant dataset, aggregation, or module.
2) **`optional`**. Parameters with default values, including `NULL` defaults that intentionally disable optional behavior.

Leave the `default` and example entries available for reference. Add new dataset and aggregation entries for your own data rather than editing target code.

## Specifying 10x reactions

Reactions are listed in `cfg_reactions.tsv`. Each row is one 10x reaction, and the `dataset` column links that reaction to a matching entry in `cfg_datasets.yaml`.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("../cfg_reactions_template.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

The columns are:

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used throughout target names and downstream config. |
| `dataset` | Dataset key used to join dataset-level settings from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor ID assigned to all nuclei in a non-multiplexed reaction. Use `NA` for multiplexed reactions. |
| `reaction_n_donors` | Number of donors expected in the reaction. |
| `reaction_cellranger_count_dir` | Path to the `cellranger-arc count` output directory. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file. Use `NA` when not used. |
| `reaction_donors_VCF_file` | Optional VCF for donor demultiplexing. Use `NA` when not used. |

Dataset-level settings live in `cfg_datasets.yaml`. This is where shared settings for a dataset are defined once and applied to all reactions that reference that dataset in `cfg_reactions.tsv`.

Typical dataset-level settings include:

- `dataset_cellranger_arc_refdata_dir`: Cell Ranger ARC reference directory for the dataset genome.
- `dataset_QC_exclude_list_per_reaction`: reaction-level QC filters applied before aggregation.

## Specifying aggregations

Aggregations are configured in `cfg_aggregations.yaml`. An aggregation is a computational analysis that combines one or more configured reactions and controls the merged RNA/ATAC processing, clustering, annotation, plotting, and optional downstream modules.

At minimum, each aggregation needs:

| Field | Meaning |
|---|---|
| `aggregation_reaction_IDs` | Reaction IDs from `cfg_reactions.tsv` to include in the joint analysis. |
| `aggregation_GEX_marker_genes` | Named marker-gene lists used for module scoring and cluster annotation. |
| `aggregation_donor_id_metadata_tsv` | Donor-level metadata table with a `donor_id` column. |
| `aggregation_reaction_ID_metadata_tsv` | Reaction-level metadata table with a `TENX_reaction_ID` column. |

Optional aggregation settings control QC filters, RNA and ATAC dimensions, clustering resolution, UMAP parameters, subgroup analysis, plotting variables, track export, and enabled modules.

For example, this field enables optional modules for an aggregation:

```yaml
optional:
  modules: [genetic_enrichment]
```

Only enable a module when the aggregation has the inputs that module needs.

## Distributed computing
The pipelines support distributed computing through the [targets](https://books.ropensci.org/targets/crew.html) framework. The active backend is configured in `crew_controllers.R`, which can be a local symlink or copy.

`crew_controllers.R` must return a named list with:

- `controller_list`: the list of `crew` controllers.
- `controller_resources_tibble`: a table with `controller_name`, `cores`, `RAM_GB`, and `gpus`.

Targets request resources by cores, RAM, and optionally GPUs through `get_tar_resources()`. The first row of `controller_resources_tibble` is the default controller, and row order controls which matching controller is preferred. GPU controllers are excluded unless `gpus_req > 0`.

For local execution, the template copied during [Installation](installation.qmd) is usually sufficient. For cluster execution, adapt `crew_controllers.R` to match your scheduler, queue names, resource limits, and worker startup requirements.

## Module configuration

Optional aggregation modules have their own config files:

- `module_differential_analyses/cfg.yaml`
- `module_genetic_enrichment/cfg.yaml`

Only edit module rows for aggregations that enable the corresponding module in `cfg_aggregations.yaml`.

## Changing the default pipeline output directory
The active workflow is a single root [targets](https://books.ropensci.org/targets/) project. The output directory can be changed via the `store` field in the [`_targets.yaml`](https://github.com/koefoeden/multiomeR/tree/main/_targets.yaml) file.
