# Inspect the demo results

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

The public demo uses `outputs/` as its results store. Other checkouts may
use a different folder, recorded under `store` in `_targets.yaml`.
The R commands below use that configuration automatically.

The store contains serialized R objects in `objects/`, file artifacts in `files/`, and requested review figures in `plots/`.

## Objects

Most intermediate and final result objects are saved automatically by `targets` under `outputs/objects` during a pipeline run. Read them with `targets::tar_read()` from a repository-root R session after the demo has completed.

```{.r filename="R"}
cell_metadata <- targets::tar_read(
  metadata_w_cell_types_tibble.WNN.immune_human_2x
)
dim(cell_metadata)
head(cell_metadata)

demo_object <- targets::tar_read(multimodal_Seurat_object.8_multimodal_QC.immune_human_2x)
demo_object
```

The metadata table describes the retained nuclei and their annotations.
WNN means *weighted nearest neighbors*: the integrated representation uses
information from both RNA and ATAC. The Seurat/Signac object is a convenient
export for further exploration; the pipeline also retains its matrices in
BPCells format on disk.

## Files

File targets also load with `targets::tar_read()`, but their value is a path rather than an in-memory result.

```{.r filename="R"}
targets::tar_read(cellranger_barcodes_tsv.healthy_PBMC_human)
targets::tar_read(aggregated_GEX_BPCells_matrix_dir.GEX.immune_human_2x)
targets::tar_read(consensus_peak_BPCells_matrix_dir.ATAC.immune_human_2x)
```

For example, `aggregated_GEX_BPCells_matrix_dir.GEX.immune_human_2x` is placed under:

```text
outputs/files/immune_human_2x/GEX/
```

Other files are grouped under `outputs/files/<scope>/`, where the scope is a
GEM well, an internal pre-aggregation QC group, or an aggregation.

## Plots

The demo command also built `categorical.UMAPs.8_multimodal_QC.immune_human_2x`.
Plots use the same scope-based layout under `outputs/plots/`, so its files are
in:

```text
outputs/plots/immune_human_2x/8_multimodal_QC/UMAPs/categorical/
```

Open `WNN_harmony_SNN_cluster_cell_type.png` there to see the integrated
clusters and cell-type labels from your own run. It should resemble this
documentation snapshot:

```{r, echo = FALSE, eval = TRUE, results = "asis"}
render_gallery_cards(gallery_items, "wnn-umap")
```

The [Main pipeline gallery](gallery_main.qmd) shows the other plot families
the workflow produces. Those are saved snapshots and do not reflect the state of
your analysis. To build one more of them, follow [Request an additional
result](main_running.qmd#request-an-additional-result); to build all review
plots for a stage, run its step in [Run your own
analysis](main_running.qmd#steps).

## Next steps

- To adopt the workflow, continue with [Plan your analysis](main_overview.qmd) and the steps in [Run your own analysis](main_running.qmd#steps).
- To request more results or rerun after a change, use [Run your own analysis](main_running.qmd).
- To diagnose a failed or unexpectedly stale target, use [Troubleshooting](troubleshooting.qmd).
