# Differential analyses module

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant when an aggregation has experimental, clinical, phenotypic, or technical variables that should be modelled across processed Multiome data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_differential_analyses/`.

The module can model:

- differential cell type composition using generalized linear mixed models
- differential gene expression on pseudobulked RNA counts
- differential chromatin accessibility on pseudobulked ATAC counts
- differential transcription factor activity using chromVAR-derived pseudobulk signals

Representative outputs from this module are shown in the [output gallery](output_gallery.qmd).

Enable the module for an aggregation in `cfg_aggregations.yaml`, then configure the module-specific model settings in `module_differential_analyses/cfg.yaml`. Only edit module rows for aggregations that enable this module.

For example, this aggregation setting enables the module:

```yaml
optional:
  modules: [differential_analyses]
```

The module config controls which variables are modelled, which metadata columns are used, and which contrasts are run. Start from `module_differential_analyses/cfg_template.yaml`, copy it to `module_differential_analyses/cfg.yaml`, and keep disabled or example rows available for reference until the corresponding aggregation is ready.

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), first complete the main pipeline checkpoints for the aggregation. Then run the module checkpoint. Replace `your_aggregation` with an aggregation that enables the module in `cfg_aggregations.yaml`.

### Differential analyses checkpoint

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:differential_analyses")) &
    tidyselect::matches("your_aggregation")
)
```

This checkpoint writes the differential cell-type composition plots, pseudobulk model diagnostics, volcano plots, GSEA summaries, and cross-modality summaries configured for the aggregation. Runtime depends on the number of modelled variables, cell types, donors, and pseudobulk contrasts.

## Implementation details

The diagram below is meant to make the module easier to reason about before reading the target code directly. It is generated from tagged target metadata and the real `{targets}` dependency graph, then simplified by pruning or bypassing lower-level nodes that would make the view harder to read. It keeps real target names and preserves the dependency structure where practical, while staying compact enough to build intuition about the main control points.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
emit_mermaid("website/figures/human_curated/differential_analyses_v2.mmd")
```
