# Differential analyses module

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant when an aggregation has experimental, clinical, phenotypic, or technical variables that should be modelled across processed Multiome data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_differential_analyses/`.

The module can model:

- differential cell type composition using generalized linear mixed models
- differential gene expression on pseudobulked RNA counts
- differential chromatin accessibility on pseudobulked ATAC counts
- differential transcription factor activity using chromVAR-derived pseudobulk signals

Representative outputs from this module are shown in the [output gallery](output_gallery.qmd).

Enable the module for an aggregation in `cfg_aggregations.yaml`, then configure the module-specific model settings in `module_differential_analyses/cfg.yaml`. See [Configuration](configuration.qmd) for the general config workflow.

{{< include helpers/_overview_snippet.qmd >}}

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_differential_analyses"
overview_heading <- "## Graph-backed blood_human draft"
overview_description <- "Graph-backed draft of module inputs, model-specific matrices, DGE/DCA/DTFA results, cell-type composition models, GSEA, and diagnostic outputs."
```

{{< include helpers/_overview_snippet.qmd >}}

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the relevant root workflow targets. Replace `your_aggregation` with an aggregation that enables the module in `cfg_aggregations.yaml`.

```{r, eval = FALSE}
targets::tar_make(names = matches("your_aggregation"))
```

Runtime depends on the number of modelled variables, cell types, donors, and pseudobulk contrasts in the aggregation.
