# Genetic Enrichment

```{r setup, include = FALSE}
pipeline_name <- "genetic_enrichment"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant for human-data aggregations where GWAS or fine-mapped trait evidence should be integrated with chromatin accessibility data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_genetic_enrichment/`.

Representative outputs from this module are shown in the [genetic enrichment output gallery](gallery_genetic_enrichment.qmd).

## Features

The genetic enrichment module links processed human multiome aggregations to fine-mapped GWAS and Open Targets evidence.

- **GWAS input handling**
  - Configured study metadata, credible sets, ancestry/sample metadata, and annotation tracks.
  - Posterior-probability weighting of credible-set variants.
  - Mapping of genetic evidence onto consensus ATAC peaks.

- **Single-nucleus enrichment**
  - [`chromVAR`](https://greenleaflab.github.io/chromVAR/articles/Introduction.html) / [`gchromVAR`](https://caleblareau.github.io/gchromVAR/reference/index.html) Z-scores for configured traits.
  - Summaries by clusters, donors, and cell types.
  - Heatmaps, boxplots, and metadata-linked review plots.

- **Graph-based enrichment**
  - [`SCAVENGE`](https://github.com/sankaranlab/SCAVENGE) trait-score propagation across PCA, LSI, and WNN graph representations.
  - TRS UMAPs, enrichment heatmaps, cluster-level boxplots, and significant-cell proportion summaries.

- **Cell-type enrichment and attribution**
  - GWAS chromVAR deviations from ATAC counts summed over each cell type, with nuclei support shown alongside the heatmap.
  - Exact decomposition of each deviation into peak, credible-variant, and locus contributions.
  - Relative-deviation heatmaps, locus waterfalls, and variant-level accessibility tracks.

## Configuration

Enable the module for a human-data aggregation in `cfg_aggregations.yaml`, then configure Open Targets studies and module-specific settings in `module_genetic_enrichment/cfg.yaml`. Only edit module rows for aggregations that enable this module.

For example, this aggregation setting enables the module:

```{.yaml filename="YAML"}
modules: [genetic_enrichment]
```

The module config controls which Open Targets studies are used, how enrichment inputs are selected, and which trait-relevance summaries are produced. The manifest-backed overview below is the reference for genetic-enrichment module parameters.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("genetic_enrichment")
```

The public example below is rendered directly from `module_genetic_enrichment/cfg_template.yaml`:

<details>
<summary>Show <code>immune_human_2x</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(module_cfg_template_file, "immune_human_2x")
```

</details>

Edit `module_genetic_enrichment/cfg.yaml`, keeping disabled or example rows available for reference until the corresponding human-data aggregation is ready.

## Running the pipeline

From the repository-root R session described in [Installation](demo_installation.qmd), first complete the main pipeline checkpoints for the human-data aggregation. Then run the module checkpoint. Replace `your_aggregation` with an aggregation that enables the module in `cfg_aggregations.yaml`.

### Genetic enrichment checkpoint

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:genetic_enrichment")) &
    tidyselect::matches("your_aggregation")
)
```

This checkpoint writes the configured cell-type, attribution, and single-nucleus trait-relevance summaries, including GWAS chromVAR and SCAVENGE outputs. Runtime depends on the number of configured GWAS studies, enrichment methods, cells, and attributed loci.
