# Differential analyses

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant when an aggregation has experimental, clinical, phenotypic, or technical variables that should be modelled across processed Multiome data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_differential_analyses/`.

Representative outputs from this module are shown in the [differential analyses output gallery](gallery_differential_analyses.qmd).

## Features

The differential analyses module extends a completed aggregation with configurable statistical models for composition and pseudobulk feature testing.

- **Differential modelling**
  - Cell-type composition models for phenotype, clinical, technical, or experimental variables.
  - Pseudobulk differential gene expression, chromatin accessibility, and transcription-factor activity.
  - Model fitting with [`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html) and [`limma`](https://bioconductor.org/packages/release/bioc/html/limma.html).

- **Configuration-driven contrasts**
  - Dynamic model branches defined from module configuration.
  - Shared donor metadata and modality-specific filtering.
  - Reusable model specifications across GEX, ATAC, and TF-activity analyses.

- **Result interpretation**
  - Pseudobulk depth summaries and P-value density diagnostics.
  - Volcano plots and significant-element summaries.
  - Cross-modality summaries comparing significant signals across data layers.

- **Biological annotation**
  - GSEA on DGE results using selected [`MSigDB`](https://www.gsea-msigdb.org/gsea/msigdb) collections and optional custom gene sets.
  - Optional links between top features and [`Open Targets`](https://platform.opentargets.org/) GWAS evidence.

## Configuration

Enable the module for an aggregation in `cfg_aggregations.yaml`, then configure the module-specific model settings in `module_differential_analyses/cfg.yaml`. Only edit module rows for aggregations that enable this module.

For example, this aggregation setting enables the module:

```{.yaml filename="YAML"}
modules: [differential_analyses]
```

The module config controls which variables are modelled, which metadata columns are used, and which contrasts are run. The manifest-backed overview below is the reference for differential-analysis module parameters.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("differential_analyses")
```

The public example below is rendered directly from `module_differential_analyses/cfg_template.yaml`:

<details>
<summary>Show <code>immune_human_2x</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(module_cfg_template_file, "immune_human_2x")
```

</details>

Edit `module_differential_analyses/cfg.yaml`, keeping disabled or example rows available for reference until the corresponding aggregation is ready.

## Running the pipeline

From the repository-root R session described in [Installation](demo_installation.qmd), first complete the main pipeline checkpoints for the aggregation. Then run the module checkpoint. Replace `your_aggregation` with an aggregation that enables the module in `cfg_aggregations.yaml`.

### Differential analyses checkpoint

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:differential_analyses")) &
    tidyselect::matches("your_aggregation")
)
```

This checkpoint writes the differential cell-type composition plots, pseudobulk model diagnostics, volcano plots, GSEA summaries, and cross-modality summaries configured for the aggregation. Runtime depends on the number of modelled variables, cell types, donors, and pseudobulk contrasts.
