# Genetic enrichment

```{r setup, include = FALSE}
pipeline_name <- "genetic_enrichment"
source("helpers/_setup.R")
```

## When to use this module

Use this optional module for a completed human aggregation when fine-mapped GWAS evidence should be integrated with chromatin accessibility. It maps published credible-set variants to consensus peaks, calculates GWAS chromVAR-style deviations with the pipeline's chromVAR/betterChromVAR machinery, propagates trait relevance with [`SCAVENGE`](https://github.com/sankaranlab/SCAVENGE), and attributes cell-type deviations back to loci and variants.

See the [output gallery](gallery_genetic_enrichment.qmd) for representative results and the [implementation graph](implementation/implementation_genetic_enrichment.html) for upstream ATAC and WNN dependencies.

## Prerequisites

Before enabling the module, confirm that:

- the aggregation is human and has passed the GEX, ATAC, and multimodal checkpoints;
- WNN metadata and graph results, consensus peaks, ATAC counts, chromVAR objects, and GEX/ATAC embeddings are available;
- each configured `sourceId` represents the intended trait and population;
- the machine can download and retain the Open Targets study and credible-set Parquet datasets; and
- the biological question justifies cell-level, graph-propagated, or cell-type-pseudobulk enrichment rather than treating those summaries as interchangeable.

## Outputs

It produces source and fine-mapping metadata, posterior-weighted variant-to-peak mappings, single-nucleus GWAS deviations, SCAVENGE trait-relevance summaries, cell-type heatmaps, and peak-, locus-, and variant-level attribution outputs.

## Configure the module

First opt a human aggregation into the module:

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  modules: [genetic_enrichment]
```

Then create a matching row directly in `module_genetic_enrichment/cfg.yaml`.

```{.yaml filename="module_genetic_enrichment/cfg.yaml"}
your_aggregation:
  genetic_enrichment_GWAS_studies:
    lymphocyte_count:
      Category: positive_control
      sourceId: GCST90002388
      finemappingMethod: auto
    published_local_study:
      Category: trait_of_interest
      sourceId: data/published_study.GRCh38.parquet
```

`sourceId` values beginning with `GCST` use the pinned Open Targets datasets.
Every other value is a local Parquet filename, resolved from the project root
and tracked as a file target. Local files must satisfy the schema enforced by
`validate_local_finemapped_GWAS_tibble()`; their study ID, fine-mapping method,
build, credible-set probability, and provenance are read from the file rather
than repeated in YAML.

The root workflow currently pins Open Targets release `26.03`. That release identifier is recorded in downstream metadata and determines the available studies, credible sets, and fine-mapping methods.

For `finemappingMethod: auto`, multiomeR selects the first available supported method in this order: `SuSie`, `SuSiE-inf`, then `PICS`. Specify a method explicitly when the method itself is part of the analysis contract; the workflow fails if that method is unavailable for the study.

### Parameter reference

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("genetic_enrichment")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_entry(module_config_file, "immune_human_2x")
```

</details>

## Run and review

Preview the selected checkpoint targets:

```{.r filename="R"}
targets::tar_manifest(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:genetic_enrichment")
  ) & tidyselect::matches("your_aggregation"),
  callr_function = NULL
)[, c("name", "description")]
```

Then run the checkpoint:

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:genetic_enrichment")
  ) & tidyselect::matches("your_aggregation")
)
```

Before interpreting trait scores, verify the resolved source release and fine-mapping method, the number of credible-set loci and variants retained, and the overlap with consensus peaks. Then compare direct deviation summaries with SCAVENGE-propagated scores and use the heatmap glyphs to identify cluster-median enrichment supported by within-grouping BH-adjusted degree-matched permutation P-values.

Runtime and disk use grow with studies, cells, graph representations, permutations, and attributed loci. The [gallery](gallery_genetic_enrichment.qmd) uses a larger aggregation with six GEM wells and is not produced by the minimal quickstart.
