# Differential analyses

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

## When to use this module

Use this optional module when a completed aggregation contains replicated donor- or sample-level experimental, clinical, phenotypic, or technical variables that should be modelled across cell-type composition or pseudobulk molecular features.

The module does not create biological replication. The donor structure, covariates, design formula, and contrasts must be defensible for the intended analysis before the workflow is run.

See the [output gallery](gallery_differential_analyses.qmd) for representative diagnostics and the [implementation graph](implementation/implementation_differential_analyses.html) for target structure.

## Prerequisites

Before enabling the module, confirm that:

- the aggregation has passed the GEX, ATAC, and multimodal checkpoints;
- WNN cell-type metadata and GEX and ATAC pseudobulk matrices are available;
- the donor metadata contains one unique row per `donor_id` and every variable used in a model;
- model variables are donor- or pseudobulk-sample-level variables, not duplicated cell-level measurements; and
- the number and distribution of donors support the specified design and contrasts.

Use `differential_analyses_extended_donor_id_metadata_tsv` when the modelling table needs variables beyond the aggregation's normal donor metadata. It must retain the same unique `donor_id` key.

## Outputs

It produces cell-type-composition models; pseudobulk differential gene expression, peak accessibility, JASPAR motif-family accessibility, and expression-derived CollecTRI TF activity; diagnostics and cross-modality summaries; and competitive gene-set tests against the MSigDB Hallmark and Reactome collections.

The DCTA branch infers signed TF or TF-complex activity from normalized GEX pseudobulks with CollecTRI regulons and the `decoupleR` univariate linear model (ULM). Genes are filtered for expression across cell-type pseudobulks, and each retained regulator must have at least five measured targets. Its inferred activities then use the same configured donor-level models and contrasts as DGE, DCA, and DTFA. The published human CollecTRI network is downloaded from the OmniPath rescue archive and accepted only when it matches the pipeline's pinned SHA-256 checksum.

DTFA tests the 233 sequence-similarity families in the official JASPAR2026 CORE vertebrate clustering rather than individual TF motifs. Each family is represented by its published root motif, which is scanned directly against the consensus peaks; individual member motifs are used only as family metadata. The same family-level accessibility matrix supports marker plots and the Seurat compatibility export. The CollecTRI-DTFA comparison maps individual CollecTRI regulators to these JASPAR families and compares model t-statistics, not raw activity scales. AP1 and NFKB remain intact as complex regulons during activity inference; their canonical members are used only to associate the complexes with motif families for comparison. Detailed source-level results retain TF expression as a third reference, while family-level summaries use the median CollecTRI regulator t-statistic and report whether any mapped source is FDR-significant.

Each gene-set collection is tested independently with `cameraPR`, `inter.gene.cor = 0.01`, and a minimum of 10 genes represented in the contrast-specific universe. A significant set is more strongly associated with the contrast than the remaining tested genes, rather than merely showing any collective change. Open Targets evidence annotation is optional.

## Configure the module

First opt the aggregation into the module:

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  modules: [differential_analyses]
```

Then create a matching row in `module_differential_analyses/cfg.yaml`. The committed file is a symlink to the public template; preserve the template as a reference and use a project-specific file for a real analysis.

```{.yaml filename="module_differential_analyses/cfg.yaml"}
your_aggregation:
  differential_analyses_DCTC_plot_phenotype_vars: condition
  differential_analyses_DCTC_formula_chr: >-
    cbind(n_nuclei, n_other_nuclei) ~ 0 + condition
  differential_analyses_psbulk_DX_models:
    condition_model:
      cell_type_subset: NULL
      design_matrix_func_name: NULL
      formula: ~ 0 + cluster + condition
      random_effect: NULL
      contrast_specs_vec:
        treated_vs_control: conditiontreated
```

The full module checkpoint requests both composition and pseudobulk review outputs. Configure the DCTC phenotype/formula and at least one pseudobulk model before using that broad selector. Formula terms and contrast coefficients must match columns produced by the model matrix.

### Parameter reference

The OLINK and bulk-RNA path fields are reserved optional integration inputs and are not consumed by the current public differential-analysis checkpoint. Leave them `NULL` unless the corresponding integration is implemented in your downstream workflow.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("differential_analyses")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(module_cfg_template_file, "immune_human_2x")
```

</details>

## Run and review

Preview the selected targets first:

```{.r filename="R"}
targets::tar_manifest(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:differential_analyses")
  ) & tidyselect::matches("your_aggregation"),
  callr_function = NULL
)[, c("name", "description")]
```

Then run the checkpoint:

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:differential_analyses")
  ) & tidyselect::matches("your_aggregation")
)
```

Review pseudobulk depths and retained donor counts before interpreting coefficients. Check model-matrix terms, P-value distributions, effect directions, and agreement or disagreement across DGE, DCA, DTFA, and DCTA. The CollecTRI-DTFA concordance target summarizes family coverage, rank correlation, directional agreement, and joint FDR support for every configured contrast. These are complementary regulatory readouts: agreement strengthens a shared interpretation, while disagreement can reflect post-transcriptional regulation, motif-family ambiguity, or different evidence carried by expression and accessibility. Treat the [gallery](gallery_differential_analyses.qmd) as a visual reference, not as a statistical acceptance threshold.

Runtime depends on donors, cell types, models, contrasts, and GSEA branches. Use [Troubleshooting](troubleshooting.qmd) if a formula, contrast, or metadata join fails.
