# Differential analyses

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

## When to use this module

Use this optional module when a completed aggregation contains replicated donor- or sample-level experimental, clinical, phenotypic, or technical variables that should be modelled across cell-type composition or pseudobulk molecular features.

The module does not create biological replication. The donor structure, covariates, design formula, and contrasts must be defensible for the intended analysis before the workflow is run.

Representative outputs are shown in the [differential analyses gallery](gallery_differential_analyses.qmd), and the target structure is shown in the [implementation graph](implementation/implementation_differential_analyses.html).

## Prerequisites

Before enabling the module, confirm that:

- the aggregation has passed the GEX, ATAC, and multimodal checkpoints;
- WNN cell-type metadata and GEX, ATAC, and TF-activity pseudobulk matrices are available;
- the donor metadata contains one unique row per `donor_id` and every variable used in a model;
- model variables are donor- or pseudobulk-sample-level variables, not duplicated cell-level measurements; and
- the number and distribution of donors support the specified design and contrasts.

Use `differential_analyses_extended_donor_id_metadata_tsv` when the modelling table needs variables beyond the aggregation's normal donor metadata. It must retain the same unique `donor_id` key.

## What it produces

- Differential cell-type-composition models and coefficient summaries.
- Pseudobulk differential gene expression, chromatin accessibility, and TF activity using [`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html) and [`limma`](https://bioconductor.org/packages/release/bioc/html/limma.html).
- Depth and P-value diagnostics, volcano plots, significant-element counts, and cross-modality summaries.
- GSEA of DGE results with selected [`MSigDB`](https://www.gsea-msigdb.org/gsea/msigdb) collections and optional custom gene sets.
- Optional Open Targets evidence annotations for top features.

## Configure the module

First opt the aggregation into the module:

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  modules: [differential_analyses]
```

Then create a matching row in `module_differential_analyses/cfg.yaml`. The committed file is a symlink to the public template; preserve the template as a reference and use a project-specific file for a real analysis.

```{.yaml filename="module_differential_analyses/cfg.yaml"}
your_aggregation:
  differential_analyses_DCTC_plot_phenotype_vars: condition
  differential_analyses_DCTC_formula_chr: >-
    cbind(n_nuclei, n_other_nuclei) ~ 0 + condition
  differential_analyses_psbulk_DX_models:
    condition_model:
      cell_type_subset: NULL
      design_matrix_func_name: NULL
      formula: ~ 0 + cluster + condition
      random_effect: NULL
      contrast_specs_vec:
        treated_vs_control: conditiontreated
```

The full module checkpoint requests both composition and pseudobulk review outputs. Configure the DCTC phenotype/formula and at least one pseudobulk model before using that broad selector. Formula terms and contrast coefficients must match columns produced by the model matrix.

### Parameter reference

The OLINK and bulk-RNA path fields are reserved optional integration inputs and are not consumed by the current public differential-analysis checkpoint. Leave them `NULL` unless the corresponding integration is implemented in your downstream workflow.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("differential_analyses")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(module_cfg_template_file, "immune_human_2x")
```

</details>

## Run and review

Preview the selected targets first:

```{.r filename="R"}
targets::tar_manifest(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:differential_analyses")
  ) & tidyselect::matches("your_aggregation"),
  callr_function = NULL
)[, c("name", "description")]
```

Then run the checkpoint:

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(
    tidyselect::contains("checkpoint:differential_analyses")
  ) & tidyselect::matches("your_aggregation")
)
```

Review pseudobulk depths and retained donor counts before interpreting coefficients. Check model-matrix terms, P-value distributions, effect directions, and agreement or disagreement across DGE, DCA, and DTFA. Treat the [gallery](gallery_differential_analyses.qmd) as a visual reference, not as a statistical acceptance threshold.

Runtime depends on the number of donors, cell types, model specifications, contrasts, and GSEA branches. Use [Troubleshooting](troubleshooting.qmd) if a formula, contrast, or metadata join fails.
