# Main pipeline gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../packages/multiomeRCore/R/null_default.R")
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

These documentation snapshots show representative outputs from the public
`immune_human_2x` configuration. The first cards summarize pre-aggregation QC;
GEX, ATAC, and WNN cards are aggregation-level. Each card names its target.

Browse these examples before installing, or use them alongside your own
results. To reproduce a plot, [install](demo_installation.qmd) the demo and
[request the target named on its card](main_running.qmd#request-an-additional-result).
The demo's final-object command does not build every gallery plot. The
[QC checkpoint procedure](main_running.qmd#qc-checkpoints) is under development.

```{r, echo = FALSE, results = "asis"}
render_gallery_section(
  gallery_items,
  "Main pipeline",
  subsection_descriptions = c(
    "Parallel GEM well pre-processing" = paste(
      "Cross-GEM-well summaries compare inputs before aggregation and help",
      "identify GEM well-specific QC problems."
    ),
    "GEX" = "Aggregation-level RNA dimension-reduction and cell-type review outputs.",
    "ATAC" = "Aggregation-level chromatin-accessibility dimension-reduction outputs.",
    "WNN" = "Integrated multimodal structure and modality-contribution outputs."
  )
)
```
