# Genetic enrichment gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

These are curated outputs from the larger `PBMC_human_6x` aggregation, not from the minimal two-reaction quickstart. Reproducing them requires the completed main aggregation, the genetic-enrichment module, the configured Open Targets studies, and the relevant checkpoint targets.

The cards currently show a selected SCAVENGE/WNN view rather than every cell-type, graph, locus, and variant attribution output described by the module. See [Genetic enrichment](downstream_genetic_enrichment.qmd) for prerequisites and interpretation guidance.

```{r, echo = FALSE, results = "asis"}
render_gallery_section(
  gallery_items,
  "Genetic enrichment module",
  subsection_descriptions = c(
    "Single-nucleus chromVAR+SCAVENGE" = paste(
      "These plots show graph-propagated SCAVENGE trait relevance scores.",
      "The heatmap summarizes scores by cell type and GWAS trait, while the",
      "UMAPs show per-cell scores across the WNN embedding."
    )
  )
)
```
