# Main pipeline gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

This curated gallery shows representative outputs from the public `immune_human_2x` configuration. The images are documentation snapshots, not files automatically installed with a local target store. Generate your own review plots with the checkpoint selectors in [Run and review the workflow](main_running.qmd).

The first cards summarize the shared `immune_human_dataset` across reactions. The GEX, ATAC, and WNN cards are aggregation-level outputs for `immune_human_2x`. Each card includes the corresponding stable `targets` name so you can connect the image to the workflow.

Start with [Install and prepare the demo](demo_installation.qmd), [run the demo endpoint](demo_running.qmd), and then [inspect the outputs](demo_outputs.qmd). Request the broader checkpoints when you want to reproduce the review-plot families shown here.

```{r, echo = FALSE, results = "asis"}
render_gallery_section(
  gallery_items,
  "Main pipeline",
  subsection_descriptions = c(
    "Parallel reaction pre-processing" = paste(
      "Dataset-level summaries compare reactions before aggregation and help",
      "identify reaction-specific QC problems."
    ),
    "GEX" = "Aggregation-level RNA dimension-reduction and cell-type review outputs.",
    "ATAC" = "Aggregation-level chromatin-accessibility dimension-reduction outputs.",
    "WNN" = "Integrated multimodal structure and modality-contribution outputs."
  )
)
```
