# Differential analyses gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

These cards are a curated subset from the public `immune_human_2x` configuration. They illustrate diagnostics, not acceptable effect sizes or significance patterns for another study. See [Differential analyses](downstream_differential_analyses.qmd) for prerequisites, models, and the checkpoint command.

The full module additionally produces expression-derived CollecTRI activity results and a CollecTRI-DTFA concordance plot. They are not shown below until stable public example assets are available.

```{r, echo = FALSE, results = "asis"}
render_gallery_section(
  gallery_items,
  "Differential analyses module",
  subsection_descriptions = c(
    "Gene expression" = "Pseudobulk differential gene-expression results across configured models and contrasts.",
    "Chromatin accessibility" = "Pseudobulk differential accessibility of consensus-peak features.",
    "Motif-family accessibility" = "Pseudobulk differential accessibility results for JASPAR motif families derived from ATAC data.",
    "Cell-type composition" = "Donor-level composition models with effect and uncertainty summaries."
  )
)
```
