# Main pipeline gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../R/output_gallery_helpers.R")
gallery_items <- check_output_gallery_assets(
  manifest_file = "output_gallery.yaml",
  gallery_root = "."
)
```

These documentation snapshots show representative outputs from the public
`immune_human_2x` configuration with its two active GEM wells. The cards follow
the numbered reviews in the adaptation guide. Each
card names the target that generated the displayed demo result; click an image
to open it at full resolution.

To reproduce these plot families, [install](demo_installation.qmd) and [run](demo_running.qmd) the demo, then request the broader [review checkpoints](main_running.qmd). The endpoint-only quickstart does not build every gallery plot; [inspect its local outputs](demo_outputs.qmd) to see the distinction.

```{r, echo = FALSE, results = "asis"}
render_gallery_section(
  gallery_items,
  "Main pipeline",
  subsection_descriptions = c(
    "Pre-aggregation QC" = "Unfiltered distributions used to choose per-GEM-well QC expressions.",
    "Aggregation input QC" = "Applied per-GEM-well filters and the cells retained for GEX.",
    "GEX" = "Aggregation-level RNA dimension-reduction and cell-type review outputs.",
    "Peak QC" = "Peak-based ATAC distributions used to choose aggregation-level QC expressions.",
    "Pre-LSI QC" = "Applied peak-based filters and the cells retained for ATAC dimensionality reduction.",
    "ATAC" = "Aggregation-level chromatin-accessibility dimension-reduction outputs.",
    "WNN" = "Integrated multimodal structure and modality-contribution outputs."
  )
)
```
