# Genetic enrichment module

```{r setup, include = FALSE}
pipeline_name <- "genetic_enrichment"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant for human-data aggregations where GWAS or fine-mapped trait evidence should be integrated with chromatin accessibility data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_genetic_enrichment/`.

The module can identify trait-relevant cell types and regulatory programs using:

- single-cell and pseudobulk enrichment scoring with gchromVAR
- graph-based trait relevance scoring with SCAVENGE

Representative outputs from this module are shown in the [output gallery](output_gallery.qmd).

Enable the module for a human-data aggregation in `cfg_aggregations.yaml`, then configure Open Targets studies and module-specific settings in `module_genetic_enrichment/cfg.yaml`. See [Configuration](configuration.qmd) for the general config workflow.

## Overview

This diagram started as a generated draft from the exported `blood_human` target graph and is now maintained as a human-curated documentation diagram.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
emit_mermaid("website/figures/human_curated/blood_human_genetic_enrichment_overview.mmd")
```

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the relevant root workflow targets. Replace `your_aggregation` with a human-data aggregation that enables the module in `cfg_aggregations.yaml`.

```{r, eval = FALSE}
targets::tar_make(names = matches("your_aggregation"))
```

Runtime depends on the number of configured GWAS studies, enrichment methods, cells, and pseudobulk contrasts.
