# Genetic enrichment module

```{r setup, include = FALSE}
pipeline_name <- "genetic_enrichment"
source("helpers/_setup.R")
```

## Purpose

This optional module is relevant for human-data aggregations where GWAS or fine-mapped trait evidence should be integrated with chromatin accessibility data. It is defined inside the root [`_targets.R`](`r pipeline_github_file`) workflow, with module-specific target fragments and configuration in `module_genetic_enrichment/`.

The module can identify trait-relevant cell types and regulatory programs using:

- single-cell and pseudobulk enrichment scoring with gchromVAR
- graph-based trait relevance scoring with SCAVENGE

Representative outputs from this module are shown in the [Output Gallery](output_gallery.qmd).

Enable the module for a human-data aggregation in `cfg_aggregations.yaml`, then configure Open Targets studies and module-specific settings in `module_genetic_enrichment/cfg.yaml`. Only edit module rows for aggregations that enable this module.

For example, this aggregation setting enables the module:

```{.yaml filename="YAML"}
optional:
  modules: [genetic_enrichment]
```

The module config controls which Open Targets studies are used, how enrichment inputs are selected, and which trait-relevance summaries are produced. Edit `module_genetic_enrichment/cfg.yaml`, keeping disabled or example rows available for reference until the corresponding human-data aggregation is ready.

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), first complete the main pipeline checkpoints for the human-data aggregation. Then run the module checkpoint. Replace `your_aggregation` with an aggregation that enables the module in `cfg_aggregations.yaml`.

### Genetic enrichment checkpoint

```{.r filename="R"}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:genetic_enrichment")) &
    tidyselect::matches("your_aggregation")
)
```

This checkpoint writes the configured pseudobulk and single-nucleus trait-relevance summaries, including gchromVAR and SCAVENGE outputs. Runtime depends on the number of configured GWAS studies, enrichment methods, cells, and pseudobulk contrasts.

## Implementation details

The diagrams below are meant to make the module easier to reason about before reading the target code directly. They are generated from tagged target metadata and the real `{targets}` dependency graph, then simplified by pruning or bypassing lower-level nodes that would make each view harder to read. They keep real target names and preserve the dependency structure where practical, while staying compact enough to build intuition about the main control points.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
```

### Pseudobulk GWAS

This view covers GWAS and pseudobulk inputs used for trait-relevance analyses.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/genetic_enrichment_pseudobulk_v2.mmd")
```

### Genetic enrichment across genes

This view covers across-gene enrichment outputs and downstream trait summaries.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/genetic_enrichment_single_nucleus_v2.mmd")
```
