# Differential analyses

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

`module_differential_analyses/targets.R` filters aggregations that enabled the module, joins their module config, attaches symbols for accepted WNN metadata and pseudobulk inputs, and maps the composition, pseudobulk, GSEA, and cross-modality target fragments. The generic pseudobulk model family is instantiated for DGE, DCA, DTFA, and expression-derived CollecTRI activity (DCTA). DCTA first converts filtered, normalized GEX pseudobulks to signed ULM scores and then reuses the same model and contrast machinery.

DTFA uses the 233 official JASPAR2026 CORE vertebrate familial root motifs as its complete feature universe. The pipeline scans those family-level profiles directly, rather than scanning individual motifs and taking the union of their peak matches.

The cross-modality fragment creates a CollecTRI-to-JASPAR family crosswalk, a detailed regulator-level table containing DCTA, DTFA, and TF-expression results, a family-level comparison table, a contrast-level concordance summary, and its plot. CollecTRI complexes remain intact in DCTA; complex-member mappings are introduced only by the comparison crosswalk.

The graph below is an orientation view. Inspect `setup_and_DCTC_targets.R`, `psbulk_DX_targets.R`, `GSEA_targets.R`, and `cross_modality_targets.R` for the complete model and plotting commands. The user-facing prerequisites and checkpoint selector are documented in [Differential analyses](../downstream_differential_analyses.html).

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/differential_analyses_v2.mmd")
```
