# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

The root `_targets.R` creates dataset, reaction, and aggregation mapping rows, then maps target fragments from `extra_targets/`. Use the diagrams to find the relevant stage, then inspect the corresponding source file for the complete command and resource declaration.

| Stage | Primary source |
|---|---|
| Reaction preprocessing | `extra_targets/per_reaction_targets.R` |
| Dataset summaries | `extra_targets/per_dataset_targets.R` |
| Aggregation setup and shared QC | `extra_targets/general_aggregation_targets.R` |
| GEX | `extra_targets/GEX_merge_and_dim_reduc_targets.R`, `extra_targets/GEX_graph_and_cluster_targets.R` |
| ATAC | `extra_targets/ATAC_targets.R` |
| WNN | `extra_targets/WNN_targets.R` |
| Compatibility export | `extra_targets/Seurat_Signac_export_targets.R` |
| Subgroups | `extra_targets/subgroups.R` |

## Parallel pre-processing

This view covers per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, barcode filtering, and handoffs into aggregation-level GEX and ATAC objects.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/parallel_v2.mmd")
```

## GEX processing

This view covers merged RNA processing, clustering, marker detection, cell type annotation, and GEX review outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/GEX_v2.mmd")
```

## ATAC processing

This view covers ATAC QC, peak calling, consensus peak construction, chromatin accessibility processing, chromVAR scoring, coverage tracks, and peak-gene links.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/ATAC_v2.mmd")
```

## WNN integration

This view covers GEX and ATAC embedding handoffs, WNN integration, modality weights, cluster comparison, and integrated metadata outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/WNN_v2.mmd")
```

## Subgroup reprocessing

This view covers optional subgroup-native GEX, ATAC, and WNN reprocessing for sufficiently large parent groups.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/full_subgroups_v2.mmd")
```

Subgroup reprocessing is configuration-dependent and should not be treated as part of the minimum main-pipeline path.
