# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Parallel pre-processing

This view covers per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, barcode filtering, and handoffs into aggregation-level GEX and ATAC objects.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/parallel_v2.mmd")
```

## GEX processing

This view covers merged RNA processing, clustering, marker detection, cell type annotation, and GEX review outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/GEX_v2.mmd")
```

## ATAC processing

This view covers ATAC QC, peak calling, consensus peak construction, chromatin accessibility processing, chromVAR scoring, coverage tracks, and peak-gene links.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/ATAC_v2.mmd")
```

## WNN integration

This view covers GEX and ATAC embedding handoffs, WNN integration, modality weights, cluster comparison, and integrated metadata outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/WNN_v2.mmd")
```

## Subgroup reprocessing

This view covers optional subgroup-native GEX, ATAC, and WNN reprocessing for sufficiently large parent groups.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/full_subgroups_v2.mmd")
```
