# Genetic enrichment

```{r setup, include = FALSE}
pipeline_name <- "genetic_enrichment"
source("helpers/_setup.R")
```

`module_genetic_enrichment/targets.R` filters enabled human aggregations, resolves one configured Open Targets study set per aggregation, and attaches symbols for WNN metadata, graphs, embeddings, consensus peaks, chromVAR state, and ATAC fragments.

The main target fragments live in `setup_targets.R`, `gchromVAR_targets.R`, `SCAVENGE_graph_targets.R`, `SCAVENGE_group_targets.R`, `GWAS_chromVAR_cell_type_targets.R`, and `GWAS_chromVAR_contribution_targets.R`. The user-facing release, method-selection, and interpretation contracts are documented in [Genetic enrichment](../downstream_genetic_enrichment.html).

## Single-nucleus and graph-based enrichment

This view covers the configured GWAS inputs, single-nucleus enrichment state, graph propagation, and downstream trait summaries. Additional cell-type contribution and locus-attribution branches may be pruned from this compact orientation view; use the manifest for the complete graph.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/genetic_enrichment_single_nucleus_v2.mmd")
```
