# Differential analyses

```{r setup, include = FALSE}
pipeline_name <- "differential_analyses"
source("helpers/_setup.R")
```

`module_differential_analyses/targets.R` filters aggregations that enabled the module, joins their module config, attaches symbols for accepted WNN metadata and pseudobulk inputs, and maps the composition, pseudobulk, GSEA, and cross-modality target fragments. The generic pseudobulk model family is instantiated for DGE, DCA, and DTFA.

DTFA uses the 233 official JASPAR2026 CORE vertebrate familial root motifs as its complete feature universe. The pipeline scans those family-level profiles directly, rather than scanning individual motifs and taking the union of their peak matches.

The graph below is an orientation view. Inspect `setup_and_DCTC_targets.R`, `psbulk_DX_targets.R`, `GSEA_targets.R`, and `cross_modality_targets.R` for the complete model and plotting commands. The user-facing prerequisites and checkpoint selector are documented in [Differential analyses](../downstream_differential_analyses.html).

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/differential_analyses_v2.mmd")
```
