# Installation

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## System requirements

- A Linux x86_64 machine with XYZ GBs of storage (for the demo workflow)
- `git`.


## Clone the repository

```{bash}
git clone https://github.com/koefoeden/multiomeR.git
cd multiomeR
```

The repository includes active config files for the public example workflow. These point to the example-data paths used by the demo, so no template-copy step is needed before the first run.

## Create and launch the pixi environment

If `pixi` is not installed yet, install it first:

```{bash}
curl -fsSL https://pixi.sh/install.sh | sh
export PATH="$HOME/.pixi/bin:$PATH"
```

Install the pixi environment from the repository root:

```{bash}
pixi install --run-post-link-scripts
```

Start an interactive R session inside the pixi environment:

```{bash}
pixi shell
R
```

Use this repository-root R session as the evaluation environment for subsequent R commands in this manual.

::: {.callout-note collapse="true"}
## Why pixi?

multiomeR depends on both R packages and command-line genomics tools. Pixi keeps those dependencies in one locked environment, so R, Bioconductor packages, and external tools such as `macs3`, `tabix`, and `cellsnp-lite` are resolved together instead of being installed manually across separate package managers.
:::

::: {.callout-note collapse="true"}
## Why is `--run-post-link-scripts` needed?

The `--run-post-link-scripts` flag is required for Bioconductor data packages such as the BSgenome genome packages. Pixi disables conda post-link scripts by default, but these packages use Bioconda's post-link scripts to materialize the actual R package directories in the pixi environment.
:::

## Installation of R packages

A small number of R packages still need GitHub versions while upstream conda packages catch up. Install them inside the R session started from the pixi shell:

```{r}
remotes::install_github(
  "bnprks/BPCells",
  subdir = "r",
  ref = "28759cdd512578b6cbe549e226e1cd52a2d2308c", # BPCells 0.3.1
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)
remotes::install_github(
  "stuart-lab/signac",
  ref = "5d66cf6c34322309d5b8aece3fa294f5531b8eee", # Signac 1.9999.4
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)
remotes::install_github(
  "plger/betterChromVAR",
  ref = "82ae1e4ada8a43c721aa2ae1c3f2c6dfc3d5a637", # betterChromVAR 0.99.41
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)
```

## Download public inputs

The public PBMC demo uses public `cellranger-arc count` outputs and the matching 10x Genomics Cell Ranger ARC reference directories. Download both from the repository root before running the demo:

```{bash}
bash example_data/download_10X_cellranger_count_data.sh
bash example_data/download_10X_cellranger_arc_refdata.sh
```

The default controller setup uses a local `crew` backend. See [Distributed computing](distributed_computing.qmd) when adapting the workflow to a cluster or changing resource routing.
