# Installation

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

multiomeR is developed for Linux systems and is managed with [pixi](https://pixi.sh/). The workflow is not distributed as a conventional installed R package; run it from the repository root so that `targets`, config files, helper code, and external command-line tools resolve together.

## Requirements

- A Linux x86_64 machine.
- `git`.
- `pixi`.
- Enough disk space for the pixi environment, Cell Ranger outputs, target store, and generated plots.
- Input data from `cellranger-arc count`, or the public example data described in the quickstart.

The pixi workspace pins R and most command-line tools in `pixi.toml` and `pixi.lock`. The active runtime currently targets R 4.5.x.

## Clone the repository

```{bash}
git clone https://github.com/koefoeden/multiomeR.git
cd multiomeR
```

## Create the pixi environment

If `pixi` is not installed yet, install it first:

```{bash}
curl -fsSL https://pixi.sh/install.sh | sh
export PATH="$HOME/.pixi/bin:$PATH"
```

Install the pixi environment from the repository root:

```{bash}
pixi install
```

The pixi environment provides R and the command-line tools used by the workflow, such as `bgzip`, `tabix`, `macs3`, `bedGraphToBigWig`, `bedToBigBed`, `cellsnp-lite`, and `vireo`.

Start an interactive R session inside the pixi environment:

```{bash}
pixi shell
R
```

Use this repository-root R session as the evaluation environment for subsequent R commands in this manual.

A small number of R packages still need GitHub versions while upstream conda packages catch up. Install them inside the R session started from the pixi shell:

```{r}
pak::pkg_install(c("bnprks/BPCells/r", "stuart-lab/signac@v2", "plger/betterChromVAR"))
```

## Create local config files

Copy the public templates before constructing the target graph:

```{bash}
cp cfg_datasets_template.yaml cfg_datasets.yaml
cp cfg_reactions_template.tsv cfg_reactions.tsv
cp cfg_aggregations_template.yaml cfg_aggregations.yaml
cp module_differential_analyses/cfg_template.yaml module_differential_analyses/cfg.yaml
cp module_genetic_enrichment/cfg_template.yaml module_genetic_enrichment/cfg.yaml
```

Edit the copied files when adapting the workflow to your own data. The templates remain available for reference.

## Configure local execution

Create a local controller file before constructing the target graph:

```{bash}
cp crew_controllers_template.R crew_controllers.R
```

The template uses a local `crew` controller. Increase workers, cores, and memory only after the example workflow succeeds on your machine.

## Smoke test

After the config files are in place, use the repository-root R session described above to check target graph construction:

```{r}
targets::tar_manifest(callr_function = NULL)
```

If this fails, fix the first missing config file, missing command, or missing package reported by the error before running `targets::tar_make()`.
