# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

multiomeR uses two linked configuration layers. A **GEM well** points to one
`cellranger-arc count` output and defines its pre-aggregation processing and QC.
An **aggregation** selects GEM wells for joint GEX, ATAC, and WNN analysis.

| Layer | Configuration | Key relationship |
|---|---|---|
| GEM well | `cfg_GEM_wells.tsv` | Aggregations refer to one or more `GEM_well_ID` values. |
| Aggregation | `cfg_aggregations.yaml` | Selects GEM wells and points to donor-level metadata. |

The committed `cfg_GEM_wells.tsv` and `cfg_aggregations.yaml` files may be
symlinks to repository-specific configuration. For a new project, start from
the corresponding templates and point the symlinks to reviewed project files.

## Start with one explicitly scoped analysis

`GEM_well_is_active` controls per-GEM-well graph construction, while aggregation
`is_active` controls aggregation graph construction. Every active aggregation
must reference active GEM wells. Before an unqualified `targets::tar_make()`,
deactivate every GEM well and aggregation you are not ready to run. During
setup, always select one aggregation explicitly.

The following minimum example shows the relationship between the two layers.
Replace the paths, identifiers, and marker genes with values appropriate for
your study.

### 1. Define one GEM well

```{.text filename="cfg_GEM_wells.tsv"}
GEM_well_ID	GEM_well_dataset	GEM_well_donor_id	GEM_well_n_donors	GEM_well_cellranger_arc_count_dir	GEM_well_add_cellbender	GEM_well_cellbender_h5_file	GEM_well_donors_VCF_file	GEM_well_cellranger_arc_reference_json	GEM_well_QC_exclude_list	GEM_well_is_active	GEM_well_multiplex_batch
your_GEM_well	your_dataset	donor_1	1	/path/to/your_GEM_well	FALSE	NA	NA	/path/to/reference.json	TSS.enrichment < 4 ;; nucleosome_signal > 4 ;; nCount_RNA < 250	TRUE	batch_1
```

`GEM_well_cellranger_arc_reference_json` must point to the `reference.json`
from the exact Cell Ranger ARC reference used to create that GEM well's output.
multiomeR checks that the JSON genome matches the feature HDF5 and rejects
aggregations whose GEM wells use different references.

`GEM_well_QC_exclude_list` contains zero or more complete R filter expressions
separated by ` ;; `. Expressions are evaluated individually against per-barcode
metadata, preserving their order and their separate exclusion reasons. An empty
field applies no pre-aggregation QC filters.

multiomeR always runs its
[validated BPCells-native AMULET implementation](implementation/algorithm_validation.html#bpcells-native-amulet)
on the existing fragment directory. There is no per-GEM-well switch because
the implementation is fast enough to be part of the standard QC path.

`GEM_well_cellranger_arc_count_dir` points to the directory containing `outs/`, not to `outs/` itself. The baseline pipeline requires:

```{.text filename="Text"}
<GEM_well_cellranger_arc_count_dir>/outs/
|-- summary.csv
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
`-- per_barcode_metrics.csv
```

If `GEM_well_donors_VCF_file` is configured, `atac_possorted_bam.bam` is also required for `cellsnp-lite`. Without a VCF, the pipeline skips genotype demultiplexing and assigns `GEM_well_donor_id` to every called nucleus. That donor ID must match the donor metadata table.

### 2. Create keyed donor metadata

The donor metadata table must contain one unique row per `donor_id`:

```{.text filename="donor_metadata.tsv"}
donor_id	condition
donor_1	control
```

Put donor-specific phenotypes and covariates in the donor table. Put library-,
run-, or batch-specific variables directly in `cfg_GEM_wells.tsv`, using a
`GEM_well_` prefix. Apart from their key columns, donor and GEM-well metadata
must not reuse column names.

The canonical GEM-well table may contain rows for multiple aggregations. Each
aggregation first subsets and orders it by `aggregation_GEM_well_IDs`, then
projects only the columns requested for SCT, Harmony, subgroup modelling, or
configured analyses. Consequently, changing an unused annotation or a GEM well
outside the aggregation does not invalidate expensive processing targets. The
complete annotation view is attached only to explicit export targets.

### 3. Define one aggregation

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  aggregation_GEM_well_IDs: [your_GEM_well]
  aggregation_donor_id_metadata_tsv: /path/to/donor_metadata.tsv
  aggregation_GEX_marker_genes:
    Cell_type_A: [GENE1, GENE2]
    Cell_type_B: [GENE3, GENE4]
  is_active: true
```

Omit `modules` for the first main-pipeline run. Add optional module names and matching module-config rows only after the main aggregation has passed its review checkpoints.

### 4. Validate before running

From the repository-root R session, construct the graph and inspect the targets created for the aggregation:

```{.r filename="R"}
manifest <- targets::tar_manifest(callr_function = NULL)

manifest |>
  dplyr::filter(stringr::str_ends(name, ".your_aggregation")) |>
  dplyr::select(name, description)
```

Manifest construction validates the YAML parameter schema, aggregation references to GEM wells, module names and rows, and controller setup. Metadata file contents are validated when their targets run. Fix manifest-time errors before calling `tar_make()`, then continue to [Running the workflow](main_running.qmd).

## Configuration reference

The searchable overviews below are generated from `cfg_pipeline_parameters.tsv`, the same manifest used for runtime defaults and validation. Use them to change a default after the minimum configuration works.

### GEM well columns

GEM well-level configuration lives in [`cfg_GEM_wells.tsv`](`r GEM_wells_github_file`).

| Column | Meaning |
|---|---|
| `GEM_well_ID` | Stable GEM well identifier used in target names and aggregation config. |
| `GEM_well_dataset` | Internal grouping key for pre-aggregation QC summaries. |
| `GEM_well_donor_id` | Donor assigned to a non-multiplexed GEM well; use `NA` when donor identities are resolved from a VCF. |
| `GEM_well_n_donors` | Expected donor count. |
| `GEM_well_cellranger_arc_count_dir` | Path to the directory containing `outs/`. |
| `GEM_well_add_cellbender` | Whether the standard-layout CellBender output should be used. |
| `GEM_well_cellbender_h5_file` | CellBender-corrected GEX H5 path when enabled; otherwise `NA`. |
| `GEM_well_donors_VCF_file` | Optional donor-genotype VCF for `cellsnp-lite` and `vireo`; otherwise `NA`. |
| `GEM_well_cellranger_arc_reference_json` | `reference.json` from the Cell Ranger ARC reference used for this GEM well. |
| `GEM_well_QC_exclude_list` | Complete per-barcode exclusion expressions separated by ` ;; `; empty applies none. |
| `GEM_well_is_active` | Whether to construct this GEM well's processing targets. |
| Other `GEM_well_*` columns | Optional annotations consumed only when requested by a downstream metadata view or export. |

### Aggregation parameters

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). Required parameters must resolve to a non-missing value after defaults and optional `inherits:` parents are applied. Optional parameters can remain `NULL` to disable the corresponding behavior.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("aggregation")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>
