# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Three configuration layers

The main pipeline is configured in three layers: 
- **datasets** define shared genome and QC settings for a number of reactions
- **reactions** point to individual cellranger-arc count outputs
- **aggregations** choose which reactions to process together.

Defaults, missing-value rules, data types, and allowed values for YAML-backed configuration live in `cfg_pipeline_parameters.tsv`. Dataset, aggregation, and module YAML files contain concrete config rows and optional row-to-row inheritance with `inherits:`.

## `cfg_datasets.yaml`

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`). A dataset is the shared context for one or more 10x reactions, typically a genome/reference combination plus dataset-wide QC rules. Reactions in `cfg_reactions.tsv` refer to these dataset keys, so define datasets before adding reaction rows.

In the public repository, `cfg_datasets.yaml`, `cfg_reactions.tsv`, and `cfg_aggregations.yaml` point to template files. Keep those public example entries as a reference, and use project-specific copies or downstream private config files for study-specific paths, metadata, and analysis settings.

Defaults and validation rules for dataset, aggregation, and module YAML parameters live in [`cfg_pipeline_parameters.tsv`](`r pipeline_parameters_github_file`). A YAML row only needs to list values that differ from the manifest defaults, plus any parameters where `allow_missing_after_inheritance` is `FALSE` and the manifest default is `NULL`.

The manifest-backed overview below is the reference for dataset parameters. Open a topic to inspect defaults, data types, allowed values, and examples.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("dataset")
```

The public human example uses the following dataset entry, rendered directly from `cfg_datasets.yaml`:

<details>
<summary>Show <code>immune_human_dataset</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(datasets_config_file, "immune_human_dataset")
```

</details>

## `cfg_reactions.tsv`

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

Each reaction should point to a [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directory. The pipeline expects the configured `reaction_cellranger_count_dir` to contain the standard `outs/` files, including:

```{.text filename="Text"}
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("cfg_reactions.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

The reaction table columns are:

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used throughout target names and downstream config. |
| `dataset` | Dataset key used to join dataset-level settings from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor ID assigned to all nuclei in a non-multiplexed reaction. Use `NA` for multiplexed reactions. |
| `reaction_n_donors` | Number of donors expected in the reaction. |
| `reaction_cellranger_count_dir` | Path to the `cellranger-arc count` output directory. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file. Use `NA` when not used. |
| `reaction_donors_VCF_file` | Optional VCF for donor demultiplexing. Use `NA` when not used. |

## `cfg_aggregations.yaml`

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). An aggregation combines one or more configured reactions and controls merged RNA/ATAC processing, clustering, annotation, plotting, and optional downstream modules.

The manifest-backed overview below is the reference for aggregation parameters. Required rows must resolve to a non-missing value after inheritance and defaults are applied; optional rows can remain `NULL` when the corresponding behavior should be disabled.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("aggregation")
```

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

The public `immune_human_2x` aggregation below is rendered directly from `cfg_aggregations.yaml`. It shows the required aggregation inputs plus the example-specific optional overrides; standard optional settings come from `cfg_pipeline_parameters.tsv`.

<details>
<summary>Show <code>immune_human_2x</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>
