# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## `cfg_datasets.yaml`

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`). A dataset is the shared context for one or more 10x reactions, typically a genome/reference combination plus dataset-wide QC rules. Reactions in `cfg_reactions.tsv` refer to these dataset keys, so define datasets before adding reaction rows.

Typical dataset-level settings include:

- `dataset_cellranger_arc_refdata_dir`: Cell Ranger ARC reference directory for the dataset genome.
- `dataset_QC_exclude_list_per_reaction`: reaction-level QC filters applied before aggregation.

The public PBMC example uses the following dataset entry, rendered directly from `cfg_datasets.yaml`:

<details>
<summary>Show <code>immune_human_dataset</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(datasets_config_file, "immune_human_dataset")
```

</details>

## `cfg_reactions.tsv`

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

Each reaction should point to a [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directory. The pipeline expects the configured `reaction_cellranger_count_dir` to contain the standard `outs/` files, including:

```{.text filename="Text"}
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("cfg_reactions.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

The reaction table columns are:

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used throughout target names and downstream config. |
| `dataset` | Dataset key used to join dataset-level settings from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor ID assigned to all nuclei in a non-multiplexed reaction. Use `NA` for multiplexed reactions. |
| `reaction_n_donors` | Number of donors expected in the reaction. |
| `reaction_cellranger_count_dir` | Path to the `cellranger-arc count` output directory. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file. Use `NA` when not used. |
| `reaction_donors_VCF_file` | Optional VCF for donor demultiplexing. Use `NA` when not used. |

## `cfg_aggregations.yaml`

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). An aggregation combines one or more configured reactions and controls merged RNA/ATAC processing, clustering, annotation, plotting, and optional downstream modules.

At minimum, each aggregation needs:

| Field | Meaning |
|---|---|
| `aggregation_reaction_IDs` | Reaction IDs from `cfg_reactions.tsv` to include in the joint analysis. |
| `aggregation_GEX_marker_genes` | Named marker-gene lists used for module scoring and cluster annotation. |
| `aggregation_donor_id_metadata_tsv` | Donor-level metadata table with a `donor_id` column. |
| `aggregation_reaction_ID_metadata_tsv` | Reaction-level metadata table with a `TENX_reaction_ID` column. |

Optional aggregation settings control QC filters, RNA and ATAC dimensions, clustering resolution, UMAP parameters, subgroup analysis, plotting variables, track export, and enabled modules. YAML entries are divided into `mandatory` parameters that must be set and `optional` parameters that are usually inherited from `default` unless the entry overrides them. `NULL` values intentionally disable optional behavior.

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

The public PBMC aggregation below is rendered directly from `cfg_aggregations.yaml`. It shows the required aggregation inputs plus the example-specific optional overrides; standard optional settings are inherited from `default`.

<details>
<summary>Show <code>immune_human_2x</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>
