# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

multiomeR uses three linked configuration layers. A **dataset** defines a shared reference and GEM well-level QC policy, a **GEM well** points to one `cellranger-arc count` output, and an **aggregation** selects GEM wells for joint GEX, ATAC, and WNN analysis.

| Layer | Configuration | Key relationship |
|---|---|---|
| Dataset | `cfg_datasets.yaml` | GEM wells refer to the dataset key. |
| GEM well | `cfg_GEM_wells.tsv` | Aggregations refer to one or more `GEM_well_ID` values. |
| Aggregation | `cfg_aggregations.yaml` | Metadata tables must cover the configured donor and GEM well IDs. |

The committed `cfg_datasets.yaml`, `cfg_GEM_wells.tsv`, and `cfg_aggregations.yaml` files are symlinks to public templates. Keep those templates as examples. For a real project, replace the symlinks with project-specific files or provide them from a downstream private repository.

## Migration to GEM well terminology

This release makes a breaking terminology transition to the 10x Genomics term **GEM well**. There are no compatibility aliases: update configuration, metadata, and any code that calls the renamed helpers together. Existing identifier values do not need to change.

| Previous name | Current name |
|---|---|
| `cfg_reactions.tsv` / `cfg_reactions_template.tsv` | `cfg_GEM_wells.tsv` / `cfg_GEM_wells_template.tsv` |
| `reaction_ID` or `TENX_reaction_ID` | `GEM_well_ID` |
| `aggregation_reaction_IDs` | `aggregation_GEM_well_IDs` |
| `aggregation_reaction_ID_metadata_tsv` | `aggregation_GEM_well_metadata_tsv` |
| `reaction_donor_id`, `reaction_n_donors`, `reaction_cellbender_h5_file`, `reaction_donors_VCF_file` | The corresponding `GEM_well_*` columns |
| `reaction_cellranger_count_dir` | `GEM_well_cellranger_arc_count_dir` |
| `per_reaction_targets.R` | `per_GEM_well_targets.R` |

The reference input is also narrowed from a directory to the metadata file the
pipeline actually reads. Replace `dataset_cellranger_arc_refdata_dir` with
`dataset_cellranger_arc_reference_json` and append `/reference.json` to existing
reference-directory values.

## Start with one explicitly scoped analysis

`is_active` defaults to `true` for every aggregation. Before an unqualified `targets::tar_make()`, either remove unavailable template rows or add `is_active: false` to every aggregation you are not ready to run. During setup, always select one aggregation explicitly.

The following minimum example shows the relationship between the three layers. Replace the paths, identifiers, and marker genes with values appropriate for your study.

### 1. Define one dataset

```{.yaml filename="cfg_datasets.yaml"}
your_dataset:
  dataset_cellranger_arc_reference_json: /path/to/reference.json
  dataset_QC_exclude_list_per_GEM_well:
    - TSS.enrichment < 4
    - nucleosome_signal > 4
    - nCount_RNA < 250
  dataset_run_amulet: false
```

`dataset_cellranger_arc_reference_json` is required and must point to the
`reference.json` from the exact Cell Ranger ARC reference used to create the
dataset outputs. The repository includes metadata for the public human and
mouse 2020-A references and the human 2024-A reference under
`reference_metadata/`; custom references can supply their own JSON without
retaining the full reference directory. multiomeR checks that the JSON genome
matches the feature HDF5 and rejects aggregations whose GEM wells use different
references.

The QC expressions are evaluated against per-barcode metadata. Starting with `dataset_run_amulet: false` avoids adding AMULET to the first data run; enable it deliberately after the basic input contract works.

### 2. Define one GEM well

```{.text filename="cfg_GEM_wells.tsv"}
GEM_well_ID	dataset	GEM_well_donor_id	GEM_well_n_donors	GEM_well_cellranger_arc_count_dir	GEM_well_cellbender_h5_file	GEM_well_donors_VCF_file
your_GEM_well	your_dataset	donor_1	1	/path/to/your_GEM_well	NA	NA
```

`GEM_well_cellranger_arc_count_dir` points to the directory containing `outs/`, not to `outs/` itself. The baseline pipeline requires:

```{.text filename="Text"}
<GEM_well_cellranger_arc_count_dir>/outs/
|-- summary.csv
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
`-- per_barcode_metrics.csv
```

If `GEM_well_donors_VCF_file` is configured, `atac_possorted_bam.bam` is also required for `cellsnp-lite`. Without a VCF, the pipeline skips genotype demultiplexing and assigns `GEM_well_donor_id` to every called nucleus. That donor ID must match the donor metadata table.

### 3. Create keyed metadata

The donor metadata table must contain one unique row per `donor_id`:

```{.text filename="donor_metadata.tsv"}
donor_id	condition
donor_1	control
```

The GEM well metadata table must contain one unique row per `GEM_well_ID`:

```{.text filename="GEM_well_metadata.tsv"}
GEM_well_ID	batch
your_GEM_well	batch_1
```

Apart from their key columns, the donor and GEM well tables must not reuse column names. Put donor-specific phenotypes and covariates in the donor table; put library-, run-, or batch-specific variables in the GEM well table.

The GEM well table may contain rows for multiple aggregations. Each aggregation
subsets it to `aggregation_GEM_well_IDs` before expensive targets consume it,
so metadata changes for unrelated GEM wells do not propagate downstream.

### 4. Define one aggregation

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  aggregation_GEM_well_IDs: [your_GEM_well]
  aggregation_donor_id_metadata_tsv: /path/to/donor_metadata.tsv
  aggregation_GEM_well_metadata_tsv: /path/to/GEM_well_metadata.tsv
  aggregation_GEX_marker_genes:
    Cell_type_A: [GENE1, GENE2]
    Cell_type_B: [GENE3, GENE4]
  is_active: true
```

Omit `modules` for the first main-pipeline run. Add optional module names and matching module-config rows only after the main aggregation has passed its review checkpoints.

### 5. Validate before running

From the repository-root R session, construct the graph and inspect the targets created for the aggregation:

```{.r filename="R"}
manifest <- targets::tar_manifest(callr_function = NULL)

manifest |>
  dplyr::filter(stringr::str_ends(name, ".your_aggregation")) |>
  dplyr::select(name, description)
```

Manifest construction validates the YAML parameter schema, aggregation references to GEM wells, module names and rows, and controller setup. Metadata file contents are validated when their targets run. Fix manifest-time errors before calling `tar_make()`, then continue to [Running the workflow](main_running.qmd).

## Configuration reference

The searchable overviews below are generated from `cfg_pipeline_parameters.tsv`, the same manifest used for runtime defaults and validation. Use them to change a default after the minimum configuration works.

### Dataset parameters

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`).

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("dataset")
```

<details>
<summary>Show the public <code>immune_human_dataset</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(datasets_config_file, "immune_human_dataset")
```

</details>

### GEM well columns

GEM well-level configuration lives in [`cfg_GEM_wells.tsv`](`r GEM_wells_github_file`).

| Column | Meaning |
|---|---|
| `GEM_well_ID` | Stable GEM well identifier used in target names and aggregation config. |
| `dataset` | Dataset key from `cfg_datasets.yaml`. |
| `GEM_well_donor_id` | Donor assigned to a non-multiplexed GEM well; use `NA` when donor identities are resolved from a VCF. |
| `GEM_well_n_donors` | Expected donor count. |
| `GEM_well_cellranger_arc_count_dir` | Path to the directory containing `outs/`. |
| `GEM_well_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file; otherwise `NA`. |
| `GEM_well_donors_VCF_file` | Optional donor-genotype VCF for `cellsnp-lite` and `vireo`; otherwise `NA`. |

<details>
<summary>Show the public GEM well table</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
GEM_well_template <- readr::read_tsv("cfg_GEM_wells.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(GEM_well_template, format = "html"))
cat("\n</div>\n")
```

</details>

### Aggregation parameters

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). Required parameters must resolve to a non-missing value after defaults and optional `inherits:` parents are applied. Optional parameters can remain `NULL` to disable the corresponding behavior.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("aggregation")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>
