# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

multiomeR uses three linked configuration layers. A **dataset** defines a shared reference and reaction-level QC policy, a **reaction** points to one `cellranger-arc count` output, and an **aggregation** selects reactions for joint GEX, ATAC, and WNN analysis.

| Layer | Configuration | Key relationship |
|---|---|---|
| Dataset | `cfg_datasets.yaml` | Reactions refer to the dataset key. |
| Reaction | `cfg_reactions.tsv` | Aggregations refer to one or more `reaction_ID` values. |
| Aggregation | `cfg_aggregations.yaml` | Metadata tables must cover the configured donor and reaction IDs. |

The committed `cfg_datasets.yaml`, `cfg_reactions.tsv`, and `cfg_aggregations.yaml` files are symlinks to public templates. Keep those templates as examples. For a real project, replace the symlinks with project-specific files or provide them from a downstream private repository.

## Start with one explicitly scoped analysis

`is_active` defaults to `true` for every aggregation. Before an unqualified `targets::tar_make()`, either remove unavailable template rows or add `is_active: false` to every aggregation you are not ready to run. During setup, always select one aggregation explicitly.

The following minimum example shows the relationship between the three layers. Replace the paths, identifiers, and marker genes with values appropriate for your study.

### 1. Define one dataset

```{.yaml filename="cfg_datasets.yaml"}
your_dataset:
  dataset_cellranger_arc_refdata_dir: /path/to/refdata-cellranger-arc
  dataset_QC_exclude_list_per_reaction:
    - TSS.enrichment < 4
    - nucleosome_signal > 4
    - nCount_RNA < 250
  dataset_run_amulet: false
```

`dataset_cellranger_arc_refdata_dir` is required. The QC expressions are evaluated against per-barcode metadata. Starting with `dataset_run_amulet: false` avoids adding AMULET to the first data run; enable it deliberately after the basic input contract works.

### 2. Define one reaction

```{.text filename="cfg_reactions.tsv"}
reaction_ID	dataset	reaction_donor_id	reaction_n_donors	reaction_cellranger_count_dir	reaction_cellbender_h5_file	reaction_donors_VCF_file
your_reaction	your_dataset	donor_1	1	/path/to/your_reaction	NA	NA
```

`reaction_cellranger_count_dir` points to the directory containing `outs/`, not to `outs/` itself. The baseline pipeline requires:

```{.text filename="Text"}
<reaction_cellranger_count_dir>/outs/
|-- summary.csv
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
`-- per_barcode_metrics.csv
```

If `reaction_donors_VCF_file` is configured, `atac_possorted_bam.bam` is also required for `cellsnp-lite`. Without a VCF, the pipeline skips genotype demultiplexing and assigns `reaction_donor_id` to every called nucleus. That donor ID must match the donor metadata table.

### 3. Create keyed metadata

The donor metadata table must contain one unique row per `donor_id`:

```{.text filename="donor_metadata.tsv"}
donor_id	condition
donor_1	control
```

The reaction metadata table must contain one unique row per `TENX_reaction_ID`:

```{.text filename="reaction_metadata.tsv"}
TENX_reaction_ID	batch
your_reaction	batch_1
```

Apart from their key columns, the donor and reaction tables must not reuse column names. Put donor-specific phenotypes and covariates in the donor table; put library-, run-, or batch-specific variables in the reaction table.

### 4. Define one aggregation

```{.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  aggregation_reaction_IDs: [your_reaction]
  aggregation_donor_id_metadata_tsv: /path/to/donor_metadata.tsv
  aggregation_reaction_ID_metadata_tsv: /path/to/reaction_metadata.tsv
  aggregation_GEX_marker_genes:
    Cell_type_A: [GENE1, GENE2]
    Cell_type_B: [GENE3, GENE4]
  is_active: true
```

Omit `modules` for the first main-pipeline run. Add optional module names and matching module-config rows only after the main aggregation has passed its review checkpoints.

### 5. Validate before running

From the repository-root R session, construct the graph and inspect the targets created for the aggregation:

```{.r filename="R"}
manifest <- targets::tar_manifest(callr_function = NULL)

manifest |>
  dplyr::filter(stringr::str_ends(name, ".your_aggregation")) |>
  dplyr::select(name, description)
```

Manifest construction validates the YAML parameter schema, aggregation-to-reaction references, module names and rows, and controller setup. Metadata file contents are validated when their targets run. Fix manifest-time errors before calling `tar_make()`, then continue to [Running the workflow](main_running.qmd).

## Configuration reference

The searchable overviews below are generated from `cfg_pipeline_parameters.tsv`, the same manifest used for runtime defaults and validation. Use them to change a default after the minimum configuration works.

### Dataset parameters

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`).

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("dataset")
```

<details>
<summary>Show the public <code>immune_human_dataset</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(datasets_config_file, "immune_human_dataset")
```

</details>

### Reaction columns

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`).

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used in target names and aggregation config. |
| `dataset` | Dataset key from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor assigned to a non-multiplexed reaction; use `NA` when donor identities are resolved from a VCF. |
| `reaction_n_donors` | Expected donor count. |
| `reaction_cellranger_count_dir` | Path to the directory containing `outs/`. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file; otherwise `NA`. |
| `reaction_donors_VCF_file` | Optional donor-genotype VCF for `cellsnp-lite` and `vireo`; otherwise `NA`. |

<details>
<summary>Show the public reaction table</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("cfg_reactions.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

</details>

### Aggregation parameters

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). Required parameters must resolve to a non-missing value after defaults and optional `inherits:` parents are applied. Optional parameters can remain `NULL` to disable the corresponding behavior.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("aggregation")
```

<details>
<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>
