# Configuration and inputs

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

multiomeR uses two linked configuration layers. A **GEM well** points to one `cellranger-arc count` output and defines its pre-aggregation processing and QC. An **aggregation** selects GEM wells for joint GEX, ATAC, and WNN analysis.

| Layer | Configuration | Key relationship |
|------------------------|------------------------|------------------------|
| GEM well | `cfg_GEM_wells.tsv` | Aggregations refer to one or more `GEM_well_ID` values. |
| Aggregation | `cfg_aggregations.yaml` | Selects GEM wells and points to donor-level metadata. |

The committed `cfg_GEM_wells.tsv` and `cfg_aggregations.yaml` files are the
active configuration files. They enable only the two GEM wells and the
`immune_human_2x` aggregation used by the public quickstart. Edit these files
directly when configuring another project.

## Start with one explicitly scoped analysis

`GEM_well_is_active` controls per-GEM-well graph construction, while aggregation
`is_active` controls aggregation graph construction. Every active aggregation
must reference active GEM wells. Before an unqualified `targets::tar_make()`,
deactivate every GEM well and aggregation you are not ready to run. The
committed quickstart configuration already follows this rule.

The example below describes one non-multiplexed GEM well from one donor.
Replace the paths, identifiers, and marker genes for your study. It illustrates
how the files connect; one donor is not enough for a replicated comparison.

### 1. Define one GEM well

Add a row to `cfg_GEM_wells.tsv` with these values. This vertical view is a
reading aid; the saved TSV has one GEM well per row.

| Column | Example value |
|---|---|
| `GEM_well_ID` | `your_GEM_well` |
| `GEM_well_dataset` | `your_dataset` |
| `GEM_well_donor_id` | `donor_1` |
| `GEM_well_n_donors` | `1` |
| `GEM_well_cellranger_arc_count_dir` | `/path/to/your_GEM_well` |
| `GEM_well_cellranger_arc_reference_json` | `/path/to/reference.json` |
| `GEM_well_add_cellbender` | `FALSE` |
| `GEM_well_cellbender_h5_file` | `NA` |
| `GEM_well_donors_VCF_file` | `NA` |
| `GEM_well_is_active` | `TRUE` |
| `GEM_well_multiplex_batch` | `batch_1` |

Set `GEM_well_QC_exclude_list` to the exclusions chosen for your data. The
[checkpoint guidance](main_running.qmd#qc-checkpoints) is under development;
do not treat the example's numerical cutoffs as recommendations for your tissue.

`GEM_well_cellranger_arc_reference_json` must point to the `reference.json` from the exact Cell Ranger ARC reference used to create that GEM well's output. multiomeR checks that the JSON genome matches the feature HDF5 and rejects aggregations whose GEM wells use different references.

`GEM_well_QC_exclude_list` contains zero or more complete R filter expressions separated by `;;`. Expressions are evaluated individually against per-barcode metadata, preserving their order and their separate exclusion reasons. An empty field applies no pre-aggregation QC filters.

AMULET doublet detection runs as part of the standard QC calculation; see the
[implementation and validation](implementation/algorithm_validation.html#bpcells-native-amulet)
for details.

`GEM_well_cellranger_arc_count_dir` points to the directory containing `outs/`, not to `outs/` itself. The baseline pipeline requires:

``` {.text filename="Text"}
<GEM_well_cellranger_arc_count_dir>/outs/
|-- summary.csv
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
`-- per_barcode_metrics.csv
```

If `GEM_well_donors_VCF_file` is configured, `atac_possorted_bam.bam` is also required for `cellsnp-lite`. Without a VCF, the pipeline skips genotype demultiplexing and assigns `GEM_well_donor_id` to every called nucleus. That donor ID must match the donor metadata table.

### 2. Create keyed donor metadata

The donor metadata table must contain one unique row per `donor_id`:

``` {.text filename="donor_metadata.tsv"}
donor_id	condition
donor_1	control
```

Put donor-specific phenotypes and covariates in the donor table. Put library-, run-, or batch-specific variables directly in `cfg_GEM_wells.tsv`, using a `GEM_well_` prefix. Apart from their key columns, donor and GEM-well metadata must not reuse column names.

### 3. Define one aggregation

``` {.yaml filename="cfg_aggregations.yaml"}
your_aggregation:
  aggregation_GEM_well_IDs: [your_GEM_well]
  aggregation_donor_id_metadata_tsv: /path/to/donor_metadata.tsv
  aggregation_GEX_marker_genes:
    Cell_type_A: [GENE1, GENE2]
    Cell_type_B: [GENE3, GENE4]
  is_active: true
```

Replace `GENE1`–`GENE4` with gene symbols appropriate for the tissue and
reference. Omit `modules` for the first run. After reviewing the main results,
you can enable optional analyses by adding their module names here and a
matching aggregation entry in each module's configuration.

### 4. Validate before running

From the repository-root R session, construct the graph and inspect the targets created for the aggregation:

``` {.r filename="R"}
manifest <- targets::tar_manifest(callr_function = NULL)

manifest |>
  dplyr::filter(stringr::str_ends(name, ".your_aggregation")) |>
  dplyr::select(name, description)
```

Manifest construction validates the YAML parameter schema, aggregation references to GEM wells, module names and rows, and controller setup. Metadata file contents are validated when their targets run. Fix manifest-time errors before calling `tar_make()`, then continue to [Running the workflow](main_running.qmd).

## Configuration reference

The searchable overviews below are generated from `cfg_pipeline_parameters.tsv`, the same manifest used for runtime defaults and validation. Use them to change a default after the minimum configuration works.

### GEM well columns

Use the [GEM-well example](https://github.com/koefoeden/multiomeR/blob/main/cfg_GEM_wells.tsv)
to check the full set of columns.

| Column | Meaning |
|------------------------------------|------------------------------------|
| `GEM_well_ID` | Stable GEM well identifier used in target names and aggregation config. |
| `GEM_well_dataset` | Internal grouping key for pre-aggregation QC summaries. |
| `GEM_well_donor_id` | Donor assigned to a non-multiplexed GEM well; use `NA` when donor identities are resolved from a VCF. |
| `GEM_well_n_donors` | Expected donor count. |
| `GEM_well_cellranger_arc_count_dir` | Path to the directory containing `outs/`. |
| `GEM_well_add_cellbender` | Whether the standard-layout CellBender output should be used. |
| `GEM_well_cellbender_h5_file` | CellBender-corrected GEX H5 path when enabled; otherwise `NA`. |
| `GEM_well_donors_VCF_file` | Optional donor-genotype VCF for `cellsnp-lite` and `vireo`; otherwise `NA`. |
| `GEM_well_cellranger_arc_reference_json` | `reference.json` from the Cell Ranger ARC reference used for this GEM well. |
| `GEM_well_QC_exclude_list` | Complete per-barcode exclusion expressions separated by ` ;; `; empty applies none. |
| `GEM_well_is_active` | Whether to construct this GEM well's processing targets. |
| Other `GEM_well_*` columns | Optional annotations consumed only when requested by a downstream metadata view or export. |

### Aggregation parameters

See the [aggregation example](https://github.com/koefoeden/multiomeR/blob/main/cfg_aggregations.yaml)
for complete examples. Required parameters need a value, supplied directly,
by a default, or by an `inherits:` parent. Optional parameters can remain
`NULL` where the reference below allows it.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_parameter_overview("aggregation")
```

<details>

<summary>Show the public <code>immune_human_2x</code> example</summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_entry(aggregations_config_file, "immune_human_2x")
```

</details>
