# Output Gallery

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
source("../R/output_gallery_helpers.R")
manifest_file <- "output_gallery.yaml"
gallery_items <- check_output_gallery_assets(
  manifest_file = manifest_file,
  gallery_root = "."
)
```

This gallery shows representative outputs produced by multiomeR on the public example workflow. It is meant to give a quick sense of the QC summaries, multimodal embeddings, model summaries, and enrichment plots that the pipeline can produce before you run it on your own data.

Each card includes the corresponding `targets` target name so you can connect the displayed output back to the workflow.

```{r, echo = FALSE, results = "asis"}
render_gallery_item <- function(item) {
  title <- htmltools::htmlEscape(item$title)
  description <- htmltools::htmlEscape(item$description)
  target <- htmltools::htmlEscape(item$target)
  asset <- item$asset
  asset_exists <- isTRUE(item$asset_exists)

  cat('<article class="output-gallery-card">\n')
  if (asset_exists) {
    cat(sprintf('<a href="%s"><img src="%s" alt="%s"></a>\n', asset, asset, title))
  } else {
    cat('<div class="output-gallery-placeholder">Preview pending</div>\n')
  }
  cat(sprintf('<h3>%s</h3>\n', title))
  cat(sprintf('<p>%s</p>\n', description))
  cat(sprintf('<p class="output-gallery-target"><code>%s</code></p>\n', target))
  if (asset_exists) {
    cat(sprintf('<p><a href="%s">Open full output</a></p>\n', asset))
  }
  cat('</article>\n')
}

for (section_name in unique(gallery_items$section)) {
  section_items <- dplyr::filter(gallery_items, .data$section == section_name)
  cat(sprintf("\n## %s\n\n", section_name))
  cat('<div class="output-gallery-grid">\n')
  purrr::pwalk(section_items, \(...) render_gallery_item(list(...)))
  cat('</div>\n')
}
```
