# Quickstart

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This quickstart is the shortest path through the public example workflow. It condenses the detailed [Installation](installation.qmd), [Configuration](configuration.qmd), and [Main pipeline](reaction_aggregation_walkthrough.qmd) pages into two command blocks.

The downloads include public `cellranger-arc count` outputs and matching 10x Genomics Cell Ranger ARC reference directories. They are large, and the example workflow can still take substantial time and disk space.

## 1. Prepare the example workflow

Run these commands in a terminal:

```{bash}
# Clone the repository
git clone https://github.com/koefoeden/multiomeR.git
cd multiomeR

# Install pixi and create the software environment
curl -fsSL https://pixi.sh/install.sh | sh # if pixi is not installed yet
export PATH="$HOME/.pixi/bin:$PATH"
pixi install --run-post-link-scripts

# Download public example inputs
bash example_data/download_10X_cellranger_count_data.sh
bash example_data/download_10X_cellranger_arc_refdata.sh

# Create local config files from templates
cp cfg_datasets_template.yaml cfg_datasets.yaml
cp cfg_aggregations_template.yaml cfg_aggregations.yaml
cp cfg_reactions_template.tsv cfg_reactions.tsv
cp crew_controllers_template.R crew_controllers.R
cp module_differential_analyses/cfg_template.yaml module_differential_analyses/cfg.yaml
cp module_genetic_enrichment/cfg_template.yaml module_genetic_enrichment/cfg.yaml

# Start the interactive R session used below
pixi shell
R
```

The final two commands start the interactive R session used for the next block.

## 2. Run the example workflow

Run these commands inside that R session:

```{r}
# Install GitHub-only R dependencies
remotes::install_github(
  "bnprks/BPCells",
  subdir = "r",
  ref = "28759cdd512578b6cbe549e226e1cd52a2d2308c", # BPCells 0.3.1
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)
remotes::install_github(
  "stuart-lab/signac",
  ref = "5d66cf6c34322309d5b8aece3fa294f5531b8eee", # Signac 1.9999.4
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)
remotes::install_github(
  "plger/betterChromVAR",
  ref = "82ae1e4ada8a43c721aa2ae1c3f2c6dfc3d5a637", # betterChromVAR 0.99.41
  upgrade = "never",
  dependencies = FALSE,
  build_vignettes = FALSE
)

# Run the public example targets
targets::tar_make(names = matches("healthy_PBMC_human|blood_human|brain_mouse"))
```

After the example workflow has completed, see the [output gallery](output_gallery.qmd) for representative outputs.
