# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Purpose

This workflow is defined in [`_targets.R`](`r pipeline_github_file`) and represents the main analysis pipeline after Cell Ranger output generation. It processes individual 10x reactions in parallel, combines selected reactions into aggregations, and builds RNA/ATAC outputs for clustering, cell typing, WNN integration, subgroup reprocessing, and optional downstream modules.

Representative outputs from the public example workflow are collected in the [output gallery](output_gallery.qmd).

## Inputs

The main pipeline starts from one or more [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directories. Each configured reaction should point to a directory with the expected `outs/` files, including:

```text
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Start from `cfg_reactions_template.tsv`, then add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("cfg_reactions_template.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

The reaction table columns are:

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used throughout target names and downstream config. |
| `dataset` | Dataset key used to join dataset-level settings from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor ID assigned to all nuclei in a non-multiplexed reaction. Use `NA` for multiplexed reactions. |
| `reaction_n_donors` | Number of donors expected in the reaction. |
| `reaction_cellranger_count_dir` | Path to the `cellranger-arc count` output directory. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file. Use `NA` when not used. |
| `reaction_donors_VCF_file` | Optional VCF for donor demultiplexing. Use `NA` when not used. |

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`). This is where shared settings for a dataset are defined once and applied to all reactions that reference that dataset in `cfg_reactions.tsv`.

Typical dataset-level settings include:

- `dataset_cellranger_arc_refdata_dir`: Cell Ranger ARC reference directory for the dataset genome.
- `dataset_QC_exclude_list_per_reaction`: reaction-level QC filters applied before aggregation.

Aggregation-level settings live in [`cfg_aggregations.yaml`](https://github.com/koefoeden/multiomeR/tree/main/cfg_aggregations_template.yaml). An aggregation combines one or more configured reactions and controls merged RNA/ATAC processing, clustering, annotation, plotting, and optional downstream modules.

At minimum, each aggregation needs:

| Field | Meaning |
|---|---|
| `aggregation_reaction_IDs` | Reaction IDs from `cfg_reactions.tsv` to include in the joint analysis. |
| `aggregation_GEX_marker_genes` | Named marker-gene lists used for module scoring and cluster annotation. |
| `aggregation_donor_id_metadata_tsv` | Donor-level metadata table with a `donor_id` column. |
| `aggregation_reaction_ID_metadata_tsv` | Reaction-level metadata table with a `TENX_reaction_ID` column. |

Optional aggregation settings control QC filters, RNA and ATAC dimensions, clustering resolution, UMAP parameters, subgroup analysis, plotting variables, track export, and enabled modules. YAML entries are divided into `mandatory` parameters that must be set and `optional` parameters with defaults, including `NULL` defaults that intentionally disable optional behavior.

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

## Workflow stages

The main pipeline runs these stages:

- per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, and barcode filtering
- aggregation of selected reactions into dataset-level analysis objects
- RNA processing, clustering, marker detection, and cell type annotation
- ATAC peak calling, consensus peak construction, chromatin accessibility processing, and chromVAR scoring
- WNN integration across RNA and ATAC modalities
- optional subgroup reprocessing for sufficiently large parent groups

## Overview

The diagrams below started as generated drafts from the exported `blood_human` target graph and are now maintained as human-curated documentation diagrams.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
```

### Parallel pre-processing

Reaction-level inputs, QC summaries, and handoffs into aggregation.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/parallel_v2.mmd")
```

### GEX processing

Merged GEX processing, clustering, cell typing, and GEX outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/GEX_v2.mmd")
```

### ATAC processing

ATAC QC, peak calling, consensus peaks, derived modalities, coverage tracks, and peak-gene links.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/ATAC_v2.mmd")
```

### WNN integration

GEX and ATAC embeddings, WNN integration, and integrated metadata outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/WNN_v2.mmd")
```

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the main workflow in checkpoint-sized pieces. Replace `your_aggregation` with the aggregation name from `cfg_aggregations.yaml`, for example `blood_human`.

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:GEX")) &
    tidyselect::matches("your_aggregation")
)
```

At the GEX checkpoint, inspect the aggregation-level barcode exclusion summaries, RNA-only QC plots, marker summaries, and `GEX_Seurat_object.your_aggregation`. If the GEX filtering, clustering, and annotation look acceptable, continue to the multimodal checkpoint:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:multimodal")) &
    tidyselect::matches("your_aggregation")
)
```

At the multimodal checkpoint, inspect the ATAC QC plots, peak and motif summaries, WNN UMAPs, modality-weight plots, and `multimodal_Seurat_object.your_aggregation`. If the aggregation enables the optional downstream modules, continue with their checkpoint outputs:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:differential_analyses")) &
    tidyselect::matches("your_aggregation")
)

targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:genetic_enrichment")) &
    tidyselect::matches("your_aggregation")
)
```

To run all configured aggregations for a checkpoint, omit the `& tidyselect::matches("your_aggregation")` part. To run the full workflow after checkpoint review, run:

```{r, eval = FALSE}
targets::tar_make()
```

The full main pipeline can be computationally expensive. Start with the public example or a narrow target selection, then scale to larger datasets after local paths, metadata files, and controller settings are working.

The workflow output directory can be changed with the `store` field in [`_targets.yaml`](https://github.com/koefoeden/multiomeR/tree/main/_targets.yaml).
