# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Purpose

This workflow is defined in [`_targets.R`](`r pipeline_github_file`) and represents the main analysis pipeline after Cell Ranger output generation. It processes individual 10x reactions in parallel, combines selected reactions into aggregations, and builds RNA/ATAC outputs for clustering, cell typing, WNN integration, subgroup reprocessing, and optional downstream modules.

Representative outputs from the public example workflow are collected in the [Output Gallery](output_gallery.qmd).

## Inputs

The main pipeline is configured in three layers: datasets define shared genome and QC settings, reactions point to individual 10x outputs, and aggregations choose which reactions to process together.

Each YAML file has a `default` entry with shared parameters. Named datasets and aggregations inherit from `default` unless they declare another `inherits` parent, so template entries usually only list required values and settings that intentionally differ from the defaults.

### `cfg_datasets.yaml`

Dataset-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`). A dataset is the shared context for one or more 10x reactions, typically a genome/reference combination plus dataset-wide QC rules. Reactions in `cfg_reactions.tsv` refer to these dataset keys, so define datasets before adding reaction rows.

Typical dataset-level settings include:

- `dataset_cellranger_arc_refdata_dir`: Cell Ranger ARC reference directory for the dataset genome.
- `dataset_QC_exclude_list_per_reaction`: reaction-level QC filters applied before aggregation.

The public PBMC example uses the following dataset entry, rendered directly from `cfg_datasets.yaml`:

<details>
<summary>Show <code>immune_human_dataset</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(datasets_config_file, "immune_human_dataset")
```

</details>

### `cfg_reactions.tsv`

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

Each reaction should point to a [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directory. The pipeline expects the configured `reaction_cellranger_count_dir` to contain the standard `outs/` files, including:

```text
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
reaction_template <- readr::read_tsv("cfg_reactions.tsv", show_col_types = FALSE)
cat('<div class="scrollable-table">\n')
cat(knitr::kable(reaction_template, format = "html"))
cat("\n</div>\n")
```

The reaction table columns are:

| Column | Meaning |
|---|---|
| `reaction_ID` | Stable reaction identifier used throughout target names and downstream config. |
| `dataset` | Dataset key used to join dataset-level settings from `cfg_datasets.yaml`. |
| `reaction_donor_id` | Donor ID assigned to all nuclei in a non-multiplexed reaction. Use `NA` for multiplexed reactions. |
| `reaction_n_donors` | Number of donors expected in the reaction. |
| `reaction_cellranger_count_dir` | Path to the `cellranger-arc count` output directory. |
| `reaction_cellbender_h5_file` | Optional CellBender-corrected GEX H5 file. Use `NA` when not used. |
| `reaction_donors_VCF_file` | Optional VCF for donor demultiplexing. Use `NA` when not used. |

### `cfg_aggregations.yaml`

Aggregation-level settings live in [`cfg_aggregations.yaml`](`r aggregations_github_file`). An aggregation combines one or more configured reactions and controls merged RNA/ATAC processing, clustering, annotation, plotting, and optional downstream modules.

At minimum, each aggregation needs:

| Field | Meaning |
|---|---|
| `aggregation_reaction_IDs` | Reaction IDs from `cfg_reactions.tsv` to include in the joint analysis. |
| `aggregation_GEX_marker_genes` | Named marker-gene lists used for module scoring and cluster annotation. |
| `aggregation_donor_id_metadata_tsv` | Donor-level metadata table with a `donor_id` column. |
| `aggregation_reaction_ID_metadata_tsv` | Reaction-level metadata table with a `TENX_reaction_ID` column. |

Optional aggregation settings control QC filters, RNA and ATAC dimensions, clustering resolution, UMAP parameters, subgroup analysis, plotting variables, track export, and enabled modules. YAML entries are divided into `mandatory` parameters that must be set and `optional` parameters that are usually inherited from `default` unless the entry overrides them. `NULL` values intentionally disable optional behavior.

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

The public PBMC aggregation below is rendered directly from `cfg_aggregations.yaml`. It shows the required aggregation inputs plus the example-specific optional overrides; standard optional settings are inherited from `default`.

<details>
<summary>Show <code>immune_human_2x</code></summary>

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_yaml_template_entry(aggregations_config_file, "immune_human_2x")
```

</details>

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the main workflow in checkpoint-sized pieces. Replace `your_aggregation` with the aggregation name from `cfg_aggregations.yaml`, for example `blood_human`.

### GEX checkpoint

Run the RNA-processing checkpoint first:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:GEX")) &
    tidyselect::matches("your_aggregation")
)
```

At this checkpoint, inspect the aggregation-level barcode exclusion summaries, RNA-only QC plots, marker summaries, and `GEX_Seurat_object.your_aggregation`. If the GEX filtering, clustering, and annotation look acceptable, continue to the ATAC checkpoint.

### ATAC checkpoint

Run the ATAC-processing checkpoint after the GEX review:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:ATAC")) &
    tidyselect::matches("your_aggregation")
)
```

At this checkpoint, inspect ATAC QC plots, LSI diagnostics, peak and motif summaries, coverage tracks, and TF or gene activity plots. If the ATAC processing looks acceptable, continue to the multimodal checkpoint.

### Multimodal checkpoint

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:multimodal")) &
    tidyselect::matches("your_aggregation")
)
```

At this checkpoint, inspect WNN UMAPs, modality-weight plots, multimodal cluster comparisons, and `multimodal_Seurat_object.your_aggregation`. Optional downstream modules are described in their own walkthroughs.

To run all configured aggregations for a checkpoint, omit the `& tidyselect::matches("your_aggregation")` part. To run the full workflow after checkpoint review, run:

```{r, eval = FALSE}
targets::tar_make()
```

The full main pipeline can be computationally expensive. Start with the public example or a narrow target selection, then scale to larger datasets after local paths, metadata files, and controller settings are working.

The workflow output directory can be changed with the `store` field in [`_targets.yaml`](https://github.com/koefoeden/multiomeR/tree/main/_targets.yaml).

## Implementation details

The diagrams below are meant to make the pipeline easier to reason about before reading the target code directly. They are generated from tagged target metadata and the real `{targets}` dependency graph, then simplified by pruning or bypassing lower-level nodes that would make the view harder to read. They keep real target names and preserve the dependency structure where practical, while staying compact enough to build intuition about the main control points.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
```

### Parallel pre-processing

This view covers per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, barcode filtering, and handoffs into aggregation-level GEX and ATAC objects.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/parallel_v2.mmd")
```

### GEX processing

This view covers merged RNA processing, clustering, marker detection, cell type annotation, and GEX review outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/GEX_v2.mmd")
```

### ATAC processing

This view covers ATAC QC, peak calling, consensus peak construction, chromatin accessibility processing, chromVAR scoring, coverage tracks, and peak-gene links.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/ATAC_v2.mmd")
```

### WNN integration

This view covers GEX and ATAC embedding handoffs, WNN integration, modality weights, cluster comparison, and integrated metadata outputs.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/WNN_v2.mmd")
```

### Subgroup reprocessing

This view covers optional subgroup-native GEX, ATAC, and WNN reprocessing for sufficiently large parent groups.

```{r, echo = FALSE, eval = TRUE, results = "asis"}
emit_mermaid("website/figures/human_curated/full_subgroups_v2.mmd")
```
