# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Purpose

This workflow is defined in [`_targets.R`](`r pipeline_github_file`) and represents the main analysis pipeline after Cell Ranger output generation. It processes individual 10x reactions in parallel, combines selected reactions into aggregations, and builds RNA/ATAC outputs for clustering, cell typing, WNN integration, subgroup reprocessing, and optional downstream modules.

Representative outputs from the public example workflow are collected in the [output gallery](output_gallery.qmd).

## Inputs

The main pipeline starts from one or more [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directories. Each configured reaction should point to a directory with the expected `outs/` files, including:

```text
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Start from `cfg_reactions_template.tsv`, then add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

Dataset-level and aggregation-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`) and [`cfg_aggregations.yaml`](https://github.com/koefoeden/multiomeR/tree/main/cfg_aggregations_template.yaml). See [Configuration](configuration.qmd) for how these files define filtering, aggregation, annotation, and optional module settings.

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

## Workflow stages

The main pipeline runs these stages:

- per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, and barcode filtering
- aggregation of selected reactions into dataset-level analysis objects
- RNA processing, clustering, marker detection, and cell type annotation
- ATAC peak calling, consensus peak construction, chromatin accessibility processing, and chromVAR scoring
- WNN integration across RNA and ATAC modalities
- optional subgroup reprocessing for sufficiently large parent groups

## Overview

The diagrams below started as generated drafts from the exported `blood_human` target graph and are now maintained as human-curated documentation diagrams.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/standard_node_color_legend.mmd")
```

### Parallel pre-processing

Reaction-level inputs, QC summaries, and handoffs into aggregation.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/human_curated/blood_human_parallel_preprocessing_overview.mmd")
```

### Aggregation

Merged GEX processing, clustering, cell typing, and GEX outputs.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/human_curated/blood_human_aggregation_overview.mmd")
```

### Multimodal dimension reduction

GEX and ATAC embeddings, WNN integration, and integrated metadata outputs.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/human_curated/blood_human_multimodal_dimension_reduction_overview.mmd")
```

## ATAC modality addition

ATAC QC, peak calling, consensus peaks, derived modalities, coverage tracks, and peak-gene links.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/human_curated/blood_human_atac_derived_modalities_overview.mmd")
```

## Subgroup analysis

Subgroup selection, within-subgroup embeddings, WNN integration, and subgroup marker outputs.

```{r, echo = FALSE, results = "asis"}
emit_mermaid("website/figures/human_curated/blood_human_subgroup_analysis_overview.mmd")
```

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the main workflow in checkpoint-sized pieces. Replace `your_aggregation` with the aggregation name from `cfg_aggregations.yaml`, for example `blood_human`.

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:GEX")) &
    tidyselect::matches("your_aggregation")
)
```

At the GEX checkpoint, inspect the aggregation-level barcode exclusion summaries, RNA-only QC plots, marker summaries, and `GEX_Seurat_object.your_aggregation`. If the GEX filtering, clustering, and annotation look acceptable, continue to the multimodal checkpoint:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:multimodal")) &
    tidyselect::matches("your_aggregation")
)
```

At the multimodal checkpoint, inspect the ATAC QC plots, peak and motif summaries, WNN UMAPs, modality-weight plots, and `multimodal_Seurat_object.your_aggregation`. If the aggregation enables the optional downstream modules, continue with their checkpoint outputs:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:differential_analyses")) &
    tidyselect::matches("your_aggregation")
)

targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:genetic_enrichment")) &
    tidyselect::matches("your_aggregation")
)
```

To run all configured aggregations for a checkpoint, omit the `& tidyselect::matches("your_aggregation")` part. To run the full workflow after checkpoint review, run:

```{r, eval = FALSE}
targets::tar_make()
```

The full main pipeline can be computationally expensive. Start with the public example or a narrow target selection, then scale to larger datasets after local paths, metadata files, and controller settings are working.
