# Main pipeline

```{r setup, include = FALSE}
pipeline_name <- "processing_and_aggregation"
source("helpers/_setup.R")
```

## Purpose

This workflow is defined in [`_targets.R`](`r pipeline_github_file`) and represents the main analysis pipeline after Cell Ranger output generation. It processes individual 10x reactions in parallel, combines selected reactions into aggregations, and builds RNA/ATAC outputs for clustering, cell typing, WNN integration, subgroup reprocessing, and optional downstream modules.

Representative outputs from the public example workflow are collected in the [output gallery](output_gallery.qmd).

## Inputs

The main pipeline starts from one or more [`cellranger-arc count`](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest/analysis/running-pipelines/single-library-analysis) output directories. Each configured reaction should point to a directory with the expected `outs/` files, including:

```text
<reaction_cellranger_count_dir>/outs/
|-- raw_feature_bc_matrix.h5
|-- filtered_feature_bc_matrix.h5
|-- atac_fragments.tsv.gz
|-- atac_fragments.tsv.gz.tbi
|-- atac_possorted_bam.bam
|-- per_barcode_metrics.csv
```

Reaction-level configuration lives in [`cfg_reactions.tsv`](`r reactions_github_file`). Start from `cfg_reactions_template.tsv`, then add one row per 10x reaction with a stable `reaction_ID`, the matching `dataset`, the `reaction_cellranger_count_dir`, and any donor demultiplexing fields relevant to that reaction.

Dataset-level and aggregation-level settings live in [`cfg_datasets.yaml`](`r datasets_github_file`) and [`cfg_aggregations.yaml`](https://github.com/koefoeden/multiomeR/tree/main/cfg_aggregations_template.yaml). See [Configuration](configuration.qmd) for how these files define filtering, aggregation, annotation, and optional module settings.

Each aggregation also points to metadata files:

- **`aggregation_donor_id_metadata_tsv`**: donor-level metadata with a `donor_id` column matching the donor IDs assigned during per-reaction processing.
- **`aggregation_reaction_ID_metadata_tsv`**: reaction-level metadata with a `TENX_reaction_ID` column matching the configured reaction IDs.

These metadata tables are joined to the per-cell metadata before downstream correction, plotting, and modelling.

## Workflow stages

The main pipeline runs these stages:

- per-reaction processing and QC, including optional ambient RNA correction, donor demultiplexing, doublet detection, and barcode filtering
- aggregation of selected reactions into dataset-level analysis objects
- RNA processing, clustering, marker detection, and cell type annotation
- ATAC peak calling, consensus peak construction, chromatin accessibility processing, and chromVAR scoring
- WNN integration across RNA and ATAC modalities
- optional subgroup reprocessing for sufficiently large parent groups

```{r, include = FALSE, eval = TRUE}
overview_stem <- "reaction_aggregation"
```

{{< include helpers/_overview_snippet.qmd >}}

## Graph-backed blood_human drafts

The diagrams below are generated from the exported `blood_human` target graph. They are intended as graph-backed drafts for reviewing and refining the simplified documentation diagrams.

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_parallel_preprocessing"
overview_heading <- "### Parallel pre-processing"
overview_description <- "Graph-backed draft of reaction-level inputs, QC summaries, and handoffs into aggregation."
```

{{< include helpers/_overview_snippet.qmd >}}

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_aggregation"
overview_heading <- "### Aggregation"
overview_description <- "Graph-backed draft of merged GEX processing, clustering, cell typing, and GEX outputs."
```

{{< include helpers/_overview_snippet.qmd >}}

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_multimodal_dimension_reduction"
overview_heading <- "### Multimodal dimension reduction"
overview_description <- "Graph-backed draft of GEX and ATAC embeddings, WNN integration, and integrated metadata outputs."
```

{{< include helpers/_overview_snippet.qmd >}}

## ATAC modality addition

This diagram focuses on how the ATAC modality is added to the merged RNA object, from peak calling and consensus peak construction through ATAC processing, chromVAR, and final WNN integration.

```{r, include = FALSE, eval = TRUE}
overview_stem <- "reaction_aggregation_ATAC_modality"
overview_heading <- "### Diagram"
overview_description <- "Simplified overview diagram of ATAC modality addition and WNN integration."
```

{{< include helpers/_overview_snippet.qmd >}}

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_atac_derived_modalities"
overview_heading <- "### Graph-backed blood_human draft"
overview_description <- "Graph-backed draft of ATAC QC, peak calling, consensus peaks, derived modalities, coverage tracks, and peak-gene links."
```

{{< include helpers/_overview_snippet.qmd >}}

## Subgroup analysis

This diagram focuses on the subgroup reprocessing branch that starts from the final WNN object, selects sufficiently large parent groups, reruns within-group multimodal embedding and clustering, and generates subgroup-specific QC and marker outputs.

```{r, include = FALSE, eval = TRUE}
overview_stem <- "reaction_aggregation_subgroups"
overview_heading <- "### Diagram"
overview_description <- "Simplified overview diagram of subgroup selection, reprocessing, clustering, and subgroup-specific outputs."
```

{{< include helpers/_overview_snippet.qmd >}}

```{r, include = FALSE, eval = TRUE}
overview_stem <- "blood_human_subgroup_analysis"
overview_heading <- "### Graph-backed blood_human draft"
overview_description <- "Graph-backed draft of subgroup selection, within-subgroup embeddings, WNN integration, and subgroup marker outputs."
```

{{< include helpers/_overview_snippet.qmd >}}

## Running the pipeline

From the repository-root R session described in [Installation](installation.qmd), run the main workflow in checkpoint-sized pieces. Replace `your_aggregation` with the aggregation name from `cfg_aggregations.yaml`, for example `blood_human`.

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:GEX")) &
    tidyselect::matches("your_aggregation")
)
```

At the GEX checkpoint, inspect the aggregation-level barcode exclusion summaries, RNA-only QC plots, marker summaries, and `GEX_Seurat_object.your_aggregation`. If the GEX filtering, clustering, and annotation look acceptable, continue to the multimodal checkpoint:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:multimodal")) &
    tidyselect::matches("your_aggregation")
)
```

At the multimodal checkpoint, inspect the ATAC QC plots, peak and motif summaries, WNN UMAPs, modality-weight plots, and `multimodal_Seurat_object.your_aggregation`. If the aggregation enables the optional downstream modules, continue with their checkpoint outputs:

```{r, eval = FALSE}
targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:differential_analyses")) &
    tidyselect::matches("your_aggregation")
)

targets::tar_make(
  names = targets::tar_described_as(tidyselect::contains("checkpoint:genetic_enrichment")) &
    tidyselect::matches("your_aggregation")
)
```

To run all configured aggregations for a checkpoint, omit the `& tidyselect::matches("your_aggregation")` part. To run the full workflow after checkpoint review, run:

```{r, eval = FALSE}
targets::tar_make()
```

The full main pipeline can be computationally expensive. Start with the public example or a narrow target selection, then scale to larger datasets after local paths, metadata files, and controller settings are working.
